Role of desolvation on biomolecular liquid-liquid phase separation

Version 1

Abstract

Biomolecular condensates play essential roles in cellular organization and are implicated in diverse pathological processes. Their formation is driven by liquid-liquid phase separation (LLPS), a process that requires coordinated multistep desolvation of biomolecular chains and multivalent inter-chain interactions. Although coarse-grained (CG) models with implicit solvent are widely used to probe LLPS thermodynamics and kinetics, they typically neglect explicit desolvation energetics, limiting their accuracy and mechanistic interpretability. Here, guided by all-atom simulations and experimental measurements, we develop a CG model that incorporates residue-level desolvation terms directly into the energy function and apply it to investigate LLPS of intrinsically disordered proteins. Incorporating explicit desolvation reshapes the phase diagram, yielding improved predictions of dense-phase packing density. Strikingly, we uncover a linear relationship between the temperature gap (simulation temperature relative to the critical point) and the extent of conformational compaction accompanying the dilute-to-dense phase transition, a result further supported by theoretical analysis. We also find that desolvation barriers accelerate early-stage coarsening dynamics while slowing chain mobility within mature condensates, whereas solvent-separated contact interactions exert the opposite effects. Together, this framework enables efficient and explicit treatment of desolvation in CG simulations and reveals how desolvation energetics shape both the thermodynamic landscape and kinetic property of biomolecular LLPS.

Introduction

Biomolecular condensates formed through Liquid-liquid phase separation (LLPS) are increasingly recognized as fundamental regulators of diverse cellular processes, including stress responses Van Der Lee et al. (2014); Feric et al. (2016); Shin and Brangwynne (2017); Boeynaems et al. (2018); Burke et al. (2015); Biamonti and Vourc’h (2010); Riback et al. (2017), cellular signaling Wippich et al. (2013); Su et al. (2016), gene expression regulation Morimoto and Boerkoel (2013); Hnisz et al. (2017), and chromatin organization Strom et al. (2017). Dysregulation of condensate formation, in contrast, has been implicated in a wide range of human diseases, most notably cancer and numerous neurodegenerative disorders Molliex et al. (2015); Patel et al. (2015); Boeynaems et al. (2018); Conicella et al. (2016); Osterburg and Dötsch (2022). Intrinsically disordered proteins (IDPs), which lack stable tertiary structures and often contain multivalent interaction motifs, are key molecular drivers of LLPS Molliex et al. (2015); Uversky et al. (2015); Zhou et al. (2018); Dignon et al. (2020); Elbaum-Garfinkle et al. (2015). Understanding the molecular principles governing IDP-mediated LLPS is therefore essential for elucidating the mechanisms underlying both physiological condensate function and pathological phase transitions. A central yet incompletely understood aspect of LLPS is the multi-step desolvation of biomolecular chains as they transition from a homogeneously dispersed solution to a condensed phase. How this desolvation energetics contributes to the thermodynamics and kinetics of LLPS remains a major open question.

Because experimentally characterizing the microscopic molecular events of LLPS at sub-molecule level is still challenging for most experimental assays, a broad spectrum of computational approaches has been developed to characterize biomolecular phase separation. These range from mean-field Flory (1942); Huggins (1941) and random-phase-approximation polymer theories Lin et al. (2017); Lin and Chan (2017); McCarty et al. (2019) to diverse molecular simulation frameworks based on atomistic or coarse-grained (CG) force fields Dignon et al. (2018b,a); Davtyan et al. (2012); De Jong et al. (2013); Baul et al. (2019). Although all-atom molecular dynamics (MD) simulations can capture the microscopic processes of biomolecules with unprecedented temporal and spatial resolutions, they become computationally prohibitive when applied to biomolecular phase separation involving large number of protein chains. This challenge has motivated the development of CG models that reduce computational cost while retaining the essential physicochemical features governing phase behavior. These CG frameworks include the sticker-and-spacer paradigm and its lattice Monte-Carlo implementation LASSI Harmon et al. (2017); Ruff et al. (2021); Choi et al. (2019), the hydrophobicity-scale (HPS) family and its refinements Dignon et al. (2018b,a); Regy et al. (2021); Dannenhoffer-Lafage and Best (2021), and Mpipi, which explicitly incorporates cation-π interactions Joseph et al. (2021). Advanced multiscale models have also been pioneered to predict the coupled phase behavior of IDPs and chromatin with remarkable accuracy Espinosa et al. (2020). Higher-resolution or experimentally fine-tuned CG force fields (e.g., AWSEM-IDP Wu et al. (2018), Martini3 Zhang et al. (2023); Souza et al. (2021), COCOMO Valdes-Garcia et al. (2023); Jussupow et al. (2025), AICG2+ Li et al. (2014); Bian et al. (2024), and SIRAH Darré et al. (2015)) further capture hydrogen bonding, secondary-structure propensity, and sequence-dependent packing. Notably, the maximum entropy optimized force field (MOFF) integrates maximum entropy biasing and energy landscape theory, successfully predicting the experimental dimensions and macroscopic phase separation of both ordered and disordered proteins Latham and Zhang (2019, 2021). More recently, Tesei et al. introduced the CALVADOS model, which optimizes residue-specific interaction parameters within the HPS framework using SAXS and paramagnetic relaxation enhancement (PRE) data across a large IDP dataset Tesei et al. (2021); Tesei and Lindorff-Larsen (2022); Tesei et al. (2024), thereby improving the accuracy and transferability of CG models for predicting IDP conformations and LLPS behavior.

Despite these advances, most residue-level CG models rely on implicit solvent representations. This simplification neglects the contribution of water molecules in mediating inter-residue interactions, which can be crucial for the thermodynamics and kinetics of biomolecular folding and phase transitions Cheung et al. (2002); Liu and Chan (2005); Karanicolas and Brooks III (2002); Tarus et al. (2006); Wu et al. (2011b). In particular, the desolvation penalty associated with electrostatic and hydrophobic interactions exhibits marked differences between implicit and explicit solvent representations Salari and Chong (2012); Wu et al. (2011a). More importantly, conventional implicit-solvent CG models cannot capture the multi-step desolvation events that accompany the transition from a dilute solution to a dense condensate in LLPS. Recent theoretical and experimental studies highlight that water reorganization and solvent-mediated entropy changes are not merely passive consequences of LLPS but active driving forces that shape condensate stability, internal structure, and dynamics Mukherjee and Schäfer (2023); Mukherjee et al. (2024); Li and Hou (2023). Nevertheless, implicit-solvent CG models remain widely used because they strike an effective balance between computational efficiency and residue-specific resolution. Incorporating explicit desolvation physics into such models thus represents a particularly promising strategy for enhancing their physical realism without sacrificing tractability in LLPS simulations. Notably, residue-level desolvation terms have already been successfully employed in protein folding studies Cheung et al. (2002); Liu and Chan (2005); Ferguson et al. (2009); Chen and Chan (2014); Zhang and Chan (2010); Karanicolas and Brooks III (2002); Gasic and Cheung (2020); Camilloni et al. (2016). Within a structure-based modeling framework, Cheung et al. introduced a desolvation barrier separating the direct-contact minimum from the solvent-separated minimum in pairwise residue-residue potentials Cheung et al. (2002). Simulations demonstrated that this minimalist desolvation model captures water-expulsion events during the final stages of folding. Chan and colleagues further showed that incorporating a desolvation barrier markedly enhances folding cooperativity Liu and Chan (2005); Ferguson et al. (2009); Chen and Chan (2014); Zhang and Chan (2010). These previous successes provide a strong foundation for the explicit implementation of desolvation effect in coarse-grained models to more realistically investigate the critical contribution of desolvation in LLPS.

In this study, we advance the coarse-grained modeling of biomolecular LLPS by explicitly incorporating desolvation terms into the energy function of CG model. By performing molecular simulations, we systematically investigate how desolvation influences phase behavior, thermodynamic properties, and the microscopic dynamics underlying condensate formation as well as the internal dynamics of the equilibrated dense phase. We further optimize the desolvation parameters using all-atom MD simulations and experimental Rg measurements, yielding a refined framework for realistically modeling sequence-dependent phase behavior of biomolecules at microscopic resolution.

Results

Explicit modeling of desolvation effects in coarse-grained models

In order to elucidate the role of water molecules in mediating inter-residue interactions, we first performed all-atom MD simulations with explicit solvent for a set of amino acid analogues and computed the potential of mean force (PMF) along inter-residue distances (r). Methane, methanol, acetate ion, ammonium ion and methanamide were chosen to represent non-polar, polar, negatively charged, positively charged, and backbone-like residues, respectively. These amino acid analogues cover the key features of most amino acid types. The simulations were performed under ambient conditions (300 K and 1 atm), and the sampled snapshots were then analyzed to determine the inter-molecule distances and the corresponding PMF profiles. Additional simulation details are provided in the Methods section.

Figure 1A shows the PMF profile for the methane system. In addition to the deep potential well corresponding to the direct contact between two methane molecules (r ∼ 4 Å), there exists a shallow potential well at larger distance (r ∼ 7 Å). This shallow minimum arises from transient, solvent-separated pseudo contacts between the solute residues, hereafter referred to as solvent-separated contact. To form a direct contact between the solutes, the water molecules mediating the solvent-separated contacts need to be excluded, which leads to a desolvation barrier between the two potential wells. Similar features have been reported in previous all-atom MD simulations Liu and Chan (2005). The PMF profiles extracted based on the simulations for other amino acid analogues exhibit the same overall shape (Figure 1B and Figure 1—figure supplement 1), but the positions of the direct-contact minimum (rdc), the solvent-separated minimum (rss), and the desolvation barrier (rb) vary depending on the analogue molecules. The depths of the two potential wells (ϵ and ϵss) and the height of the desolvation barrier (ϵb) likewise differ across analogue pairs. The desolvation effect is also intrinsically modulated by solvent polarity, which can be realized experimentally by changing ionic strength or adding cosolvents. To elucidate this relationship, we performed additional simulations of the methane system in water with systematically varied polarity. For simplicity, the solvent polarity was tuned by uniformly scaling the partial charges of the hydrogen and oxygen atoms in the water molecules. This approach directly modulates the strength of solvent-solvent electrostatic interactions, which in turn dictates the stability of the hydration shell surrounding the solute. We compared the standard water model (1.0× charges) with reduced-polarity models (0.9× and 0.8× charges) (Figure 1A). The results reveal that a reduction in solvent polarity leads to a simultaneous decrease in the height of the desolvation barrier and the depth of the solvent-separated minimum. These observations underscore the importance of explicitly modeling the desolvation energetics and exploring the effects of different desolvation strengths on the thermodynamics and kinetics of protein LLPS.

Desolvation effect of inter-molecular interactions.

(A) Potential of mean force along the inter-molecule distances for the amino acid analogues from all-atom MD simulations with different water polarity. (B) Schematic diagram of desolvation effect. (C) Pair-wise energy function incorporating desolvation terms. Different curves correspond to different desolvation parameters.

In this work, we introduce desolvation terms into the HPS model developed by Dignon et al. (2018b). In this framework, each residue is simplified to a single spherical bead. The HPS energy function includes the bonded term and nonbonded term (see Materials and methods). The nonbonded term is described by the short-range pairwise interactions and long-range electrostatic interactions. To account for the desolvation effect, we added two Gaussian functions centered at rb and rss to represent the desolvation barrier and the solvent-separated contact, respectively. The resulted nonbonded energy function with desolvation effect is given by (1) U(r)=Φ(r,ϵ,λ,σ)+Gb(r,ϵb,rb,σb)−Gss(r,ϵss,rss,σss)=Φ(r,ϵ,λ,σ)+ϵbexp⁡[−(r−rb)22σb2]−ϵssexp⁡[−(r−rss)22σss2].

Here Φ(r, ϵ, λ, σ) denotes the nonbonded term in the original HPS model energy function Dignon et al. (2018b). The parameters λ and σ correspond to the hydrophobicity scale values and effective bead sizes, respectively, for the 20 amino acids.

We first employ a homopolymer protein model with a chain length of 50 residues to illustrate the desolvation effect. For each residue, we use the parameter set extracted by averaging among the intrinsically disordered proteins in the Disprot dataset Aspromonte et al. (2024), which gives an amino acid mass of mresidue = 110 amu (atomic mass unit), a hydrophobicity parameter λ = 0.640, and an effective bead size σ = 0.536 nm. The parameter ϵb governs the height of desolvation barrier associated with expelling water molecules between residues, whereas ϵss determines the depth of the local minimum corresponding to a solvent-separated contact. Together, these parameters shape the desolvation potential and modulate the energetic cost of excluding hydration water. By tuning the parameter values of ϵss and ϵb, we generate energy functions with varying degree of desolvation contributions (Figure 1C), enabling systematic investigation of how desolvation influences the phase behavior and thermodynamics of IDPs. Unless explicitly stated otherwise, all thermodynamic quantities are reported in terms of a reduced temperature, defined as T ∗ = kBT/ϵ. Here, ϵ denotes the characteristic interaction energy scale of the system, which is set to ϵ = 0.2 kcal/mol following the convention of the HPS model Dignon et al. (2018b).

Thermodynamic impact of desolvation on the phase diagram

To capture the essential physics of IDP phase separation, we employed a coarse-grained model of 125 homopolymer chains using a slab simulation protocol (Methods). As a control, we first performed molecular dynamics simulations within the HPS model framework without considering the desolvation terms across a range of temperatures (Figure 2A-C). Representative snapshots clearly illustrate the transition from a homogeneous solution to a stable condensate as the temperature decreases (Figure 2B). We extracted time-averaged density profiles along z-axis to identify coexisting dense and dilute phases and thereby constructed the binodal curve (Figure 2A,C), which reproduces phase behaviors reported in previous molecular simulation studies Dignon et al. (2018b). We then repeated the slab simulations while explicitly including desolvation energetics with varying ϵb and ϵss to investigate how desolvation reshapes the phase diagram and condensate structure.

Thermodynamic regulation and microscopic mechanisms of desolvation-mediated phase separation.

(A) Baseline phase diagram of the poly-50 system using the standard HPS model. (B) Representative simulation snapshots visualizing the transition from a stable condensate (T ∗ = 2.58) to a near-critical state (T ∗ = 2.98) and a homogeneous solution (T ∗ = 3.18). (C) Time-averaged density profiles along the z-axis identifying the coexisting dense and dilute phases. (D, E) Macroscopic phase boundaries under varying desolvation barrier heights ϵb (D) and solvent-separated potential depths ϵss (E). Insets show the monotonic dependence of Tc on the respective parameters. The lower schematics highlight the distinct thermodynamic drivers: entropic penalty (ϵb) versus enthalpic stabilization (ϵss). (F, G) Renormalized phase behavior plotted against normalized temperature T ∗/Tc for varying ϵb (F) and varying ϵss (G).

To elucidate how desolvation regulates thermodynamics of LLPS, we systematically mapped phase diagrams under varying desolvation parameters (Figure 2D and E). Increasing the desolvation barrier ϵb monotonically lowers the critical temperature Tc∗ (Figure 2D), suggesting a reduced phase separation propensity. Analysis of residue-residue radial distribution functions showed that higher ϵb suppresses density distribution in the barrier region, suggesting an entropic penalty that restricts the conformational volume available for residue-residue association (Figure 2—figure supplement 1C). By contrast, deepening the solvent-separated well ϵss elevates Tc∗ (Figure 2E), which is attributed to an enthalpic stabilization driven by water-mediated bridging as shown by the enhanced population of solvent-separated configurations (Figure 2—figure supplement 1D). These opposing effects, entropic restriction versus enthalpic gain, suggest that the desolvation potential emerges as a key thermodynamic regulator that tunes macroscopic phase behavior via the balance of conformational entropy and solvent-mediated enthalpy (lower panels of Figure 2D and E).

Because the critical temperature is an intrinsic, sequence-specific property of a given system, it is also important to examine how desolvation strength alters the overall shape of the phase diagram when Tc∗ is held constant. To do so, we rescaled the temperature axis by Tc∗, which effectively normalizes overall energetic strength and isolates desolvation-specific effects on the phase boundary. For the desolvation models with different barrier height ϵb, the normalized binodal curves nearly overlap (Figure 2F). By contrast, varying the solvent-separated well depth ϵss results in distinct deviations in the density of the condensed phase (Figure 2G). Specifically, stronger water-mediated interactions lower equilibrium density by biasing the ensemble toward configurations with larger inter-residue separations (Figure 2—figure supplement 1E, F). Thus, solvent-separated interactions directly regulate molecular packing and, by introducing hydrated configurations, prevent artificial over-compaction as typically encountered in coarse-grained models with implicit solvent. These results suggest that explicitly including desolvation enables coarse-grained models to more faithfully capture the physicochemical structure of biological condensates.

Conformational modulation by desolvation in dense and dilute phases

We next investigate how desolvation modulates the conformational distribution of IDPs across dilute and condensed phases. To quantify the global chain dimensions, we calculated the radius of gyration (Rg) as a primary structural metric. The resulting Rg distributions for protein chains in both phases are presented in Figure 3B and C. Notably, the protein chains adopt more compact conformations in the dilute phase (smaller ⟨Rg⟩), whereas those in the condensed phase exhibit more extended conformations (larger ⟨Rg⟩). This conformational expansion upon condensation reflects a transition from the intra-chain dominated interactions to the inter-chain dominated interactions as typically encountered during phase of separation of IDPs Dignon et al. (2018b); Wei et al. (2017); Hazra and Levy (2021). Within the condensed phase, protein chains preferentially form inter-chain contacts that lower the free energy of the system, which compensate for the entropic cost associated with reduced translational and conformational freedom. The Rg distributions also suggest that chain conformations in the dilute phase are much more sensitive to variations in desolvation parameters than in the condensed phase (Figure 3B and C). Consequently, the variation in the conformational difference ΔRg=Rgdense−Rgdilute arises predominantly from the conformational changes of isolated chains in the dilute phase. Physically, ΔRg captures the transition from intramolecular-dominated collapse in the dilute phase to intermolecular-mediated expansion in the dense phase, thus serving as a metric for the relative strength of inter-chain versus intra-chain interactions.

Effect of desolvation on protein conformations.

(A) Schematic illustration of the conformational distributions of the protein in the high-and low-density phases under different desolvation parameters. (B-C) Distribution of Rg with different ϵb (B) and ϵss (C) at Tsim∗=1.59. The solid and dashed lines represents the results in the condensed and dilute phases respectively. The inset illustrates the mean value of Rg as a function of desolvation parameters. (D) Correlation between temperature difference (Tc∗−Tsim∗) and the averaged Rg difference between two phases. The purple and yellow dashed lines represent linear fits to the data obtained at low simulation temperatures (Tsim∗<1.2) and high-temperature (Tsim∗>2.4) regimes respectively. The inset displays the improved linearity obtained when the temperature difference is normalized by the simulation temperature, (Tc∗−Tsim∗)/Tsim∗. (E) Correlation between the density difference and Rg difference in the two phases with varying temperatures and desolvation parameters. The plot was shown in a log-log scale.

To further elucidate the thermodynamic implications of desolvation-modulated conformations, we analyzed the relationship between the conformational shift ΔRg and the thermal distance to the critical point, defined as Tc∗−Tsim∗. Strikingly, data from all simulated systems nearly collapse onto a single curve, revealing a strong universal correlation between the magnitude of conformational change and the thermal distance to the phase transition point (R2 = 0.942, Figure 3D). This result demonstrates that the conformational response to phase separation is governed fundamentally by how far the system resides thermally from the critical point, irrespective of the specific microscopic desolvation energetics.

We rationalize this observation using the Flory-Huggins theory Huggins (1942); Flory (1942) as a bridge between the macroscopic thermal distance and the microscopic interaction strength. In this theoretical framework, the thermodynamic state is defined by the interaction parameter χ = ϵeff/(kBT), where ϵeff represents the effective pairwise energy scale specific to each desolvation model Flory (1953). By invoking the critical condition ϵeff = kBTcχc, where χc is a universal constant determined solely by the fixed chain length N, we eliminate the system-specific energy scale ϵeff.

This substitution reveals that the deviation from the critical interaction parameter scales strictly with the normalized thermal distance, [χ(Tsim) − χc] ∝ (Tc − Tsim)/Tsim. The thermodynamic driving force χ can then be related to the structural observable ΔRg. Since ΔRg captures the structural transition from a intrachain-interaction dominated state in the dilute phase to an interchain-interaction dominated state in the dense phase, we posit that this conformational shift scales linearly with the excess interaction strength, expressed as ΔRg ∝ [χ(Tsim) − χc]. Combining these relations leads to a theoretical prediction where the thermal distance is related to the conformational change scaled by the simulation temperature (2) Tc−Tsim∝TsimΔRg.

This derivation explains the approximate linearity observed in Figure 3D while also accounting for the systematic deviation caused by the temperature coefficient, which is evident from the varying slopes between low-temperature and high-temperature data points. To validate our theoretical derivation and the underlying linear response assumption, we removed this temperature factor and replotted the data against the normalized thermal distance (Tc∗−Tsim∗)/Tsim∗. As shown in the inset of Figure 3D, this operation yields a nearly perfect linear relationship (R2 = 0.986). This statistical enhancement confirms the above argument that ΔRg scales linearly with the interaction distance χ −χc, thereby establishing that the conformational response of IDPs is intrinsically dictated by the thermal distance from the critical point.

We further show the relationship between the phase density difference (Δρ = ρdense − ρdilute) and the conformational difference (ΔRg) between coexisting phases in Figure 3E. From the previously established proportionality Tc − Tsim ∝ ΔRg, one can relate the density contrast Δρ to ΔRg through the critical-scaling relation by Δρ = A(Tc − T)β (Equation (6)), where β is the critical exponent. Substituting the linear relation Tc − Tsim ∝ ΔRg into this expression yields Δρ ∝ (ΔRg)β. On logarithmic scales, a clear power-law dependence emerges (Figure 3E), with Δρ∝ΔRg0.4, which is consistent with our theoretical expectation. The fitted exponent (≈ 0.4) is in reasonable agreement with the theoretical critical exponent β = 0.325 based on 3D Ising model Rowlinson and Widom (2013). Collectively, these results reveal a fundamental coupling between microscopic conformational reorganization and macroscopic packing density during biomolecular condensation, underlying conformational reorganization as a structural signature of the phase transition.

Dynamic consequences of desolvation in phase separation

Beyond governing thermodynamic phase boundaries and conformational distribution, the desolvation potential fundamentally reshapes the dynamical landscape of IDP condensates. The interplay between inter-chain interactions and solvent exclusion not only dictates the stability of the dense phase but also modulates the transport properties within the condensate and the macroscopic kinetics of phase separation. To probe the chain mobility within the dense phase, we employed a strategy analogous to fluorescence recovery after photobleaching (FRAP) experiments following prior molecular dynamics simulation work Yamada and Takada (2023), tracking the spatiotemporal evolution of specific chains initially localized at the slab center (Figure 4A and B). For the standard HPS model without desolvation, the tagged chains achieve a uniform distribution that matches the equilibrium density profile within 50 ns (Figure 4B). This rapid relaxation confirms a liquid-like nature of the condensate. Incorporating desolvation effects significantly slows down the relaxation process. The spatial distribution of tracked chains remains biased toward the initial center position after 50 ns and requires a prolonged timescale (∼ 150 ns) to fully equilibrate. This slowing down suggests that desolvation introduces additional roughness to the energy landscape that govern molecular transport.

Desolvation-mediated modulation of diffusion and coarsening dynamics.

(A) Snapshots of spinodal decomposition and equilibrium chain diffusion. The top panels illustrate domain coarsening, while the bottom panels highlight self-diffusion of highlighted chains in the stable slab. The inset quantifies the chain mobility via TAMSD analysis. (B) Density profiles of biomolecules along the z-axis for the entire system (orange) and for the highlighted chains (red) at different time lags with and without desolvation. The solid curves represent normalized averages over 1 μs. (C) Diffusion coefficients as a function of the inter-chain desolvation strength (ϵb, ϵss) and dense-phase densities (ρdense) at different temperatures. (D) Reduced diffusion coefficient K~ versus quench depth for varying ϵb. The schematic inserts highlight the impact of energy landscape roughness on diffusion dynamics. (E) Schematic representation of the phase separation mechanism. The diagram depicts the transition from initial density fluctuations to late-stage growth, mediated by an intermediate plateau. The zoom-in view details the structural origin of the kinetic arrest arising from strong inter-chain interactions. (F, G) Kinetics of density fluctuations and domain growth. Top: Time evolution of the normalized density variance ⟨σ^ρ2⟩ The inset shows the characteristic time τ1/2. Bottom: Growth of the average slab size ⟨ξ⟩ on a logarithmic scale. The timeline bar highlights the three distinct regimes: initial fluctuation, kinetic arrest (plateau), and late-stage coarsening via Brownian Motion Coalescence (BMC).

To more rigorously quantify this mobility, we computed the time-averaged mean squared displacement (TAMSD) Metzler et al. (2014) for chains within the slab. The chain motion exhibits sub-diffusive scaling, TAMSD = K ⋅ Δtα, with a characteristic exponent α ≈ 0.91 (Figure 4A, bottom right panel). This deviation from normal diffusion (α = 1) primarily reflects the viscoelastic nature of the crowded condensed phase with multi-valent dynamic interaction network, where chain motion is hindered by molecular crowding and chain constraints. At extended lag times, the TAMSD curves naturally plateau as they approach the geometric limit imposed by the slab thickness. Notably, our observed exponent is comparable to simulation results reported for protein diffusion in three-dimensional condensates (α ≈ 0.85) Watanabe et al. (2025), consistently highlighting the restricted nature of chain dynamics within dense phase. Given the robustness of α across different interaction parameters, we utilized the generalized diffusion coefficient K to characterize the transport rates.

The modulation of the diffusion coefficient reveals a clear dependence on the interaction parameters (Figure 4C). At a constant temperature, K increases with the desolvation barrier ϵb and decreases with the well depth ϵss (Figure 4C, left panels). These seemingly divergent trends are effectively unified by the packing state of the condensate. As illustrated in Figure 4C (right panel), the diffusion coefficient is predominantly correlated with the bulk density (ρdense), accompanied by a secondary temperature modulation, where higher temperatures provide a modest increase in mobility beyond the baseline packing constraints. Consistent with the density trends established in Figure 2D and E, a higher ϵss promotes tighter packing, increasing the effective friction, whereas a higher ϵb prevents close contacts, creating greater free volume that facilitates diffusion. These results indicate that, at a fixed absolute temperature, macroscopic density serves as the primary determinant of chain mobility.

However, this density-dominated view is incomplete because it convolutes intrinsic energy landscape features with thermodynamic state shifts. Since variations in ϵb and ϵss significantly alter the critical temperature, comparisons made at a constant absolute temperature inherently involve systems positioned at different thermodynamic quench depths, defined as the normalized temperature deviation from the critical point. This highlights the distinct thermodynamic driving forces for different systems. To disentangle the intrinsic influence of the desolvation potential from these thermal effects, we renormalized our analysis. First, we scaled the simulation temperature by the critical temperature of each system and utilized the ratio Tsim∗/Tc∗ as the control parameter for quench depth, noting that a lower value corresponds to a deeper quench. This ensures the thermodynamic equivalence across different interaction parameters. Second, to factor out the inherent linear dependence of Brownian motion on temperature where the diffusion rate scales with kBT, we defined a reduced diffusion coefficient K~=K/T as detailed in the Supplementary Information. In this renormalized frame, the intrinsic signature of the energy landscape emerges clearly (Figure 4D). At the same quench depth, where thermodynamic driving forces are comparable, a higher desolvation barrier ϵb (darker blue points) yields a lower reduced diffusion rate K~. Crucially, this kinetic divergence becomes increasingly pronounced at deeper quenches (lower Tsim∗/Tc∗) but effectively vanishes near the critical point, indicating that the regulatory role of the intrinsic energy landscape amplifies as the system moves further from the critical point. Notably, the standard HPS model exhibits much faster intrinsic mobility (Figure 4 red triangles), confirming that neglecting specific desolvation energetics artificially smooths the energy landscape and leads to an overestimation of chain dynamics. The trend shown in Figure 4C and D uncovers the physical reality masked by density effects. Specifically, it reveals that while a higher ϵb loosens the macroscopic packing, it simultaneously roughens the microscopic energy landscape. The desolvation barriers increase the activation energy required for local chain rearrangement, thereby suppressing the intrinsic mobility. Thus, the chain dynamics are governed by a hierarchy of mechanisms: macroscopic density sets the baseline mobility through free volume, while the roughness of the desolvation-mediated energy landscape imposes a fine-tuned kinetic constraint.

Desolvation modulated kinetic arrest of coarsening dynamics

Elucidating the molecular events during the early stage of phase separation is challenging experimentally. We therefore examine droplet-growth dynamics at the early stage of phase separation by performing molecular simulations with explicit consideration of desolvation effect. Initially, the system was equilibrated at a supercritical temperature (T ∗ = 2.98) for 10 ns to ensure a homogeneous distribution. The temperature was then quenched to T ∗ = 1.79, located deep within the two-phase coexisting region, to initiate spontaneous phase separation (Figure 4A). Visually, the system evolves from an initially uniform distribution into interconnected structures within 1 ns ∼ 2 ns, which subsequently breaks up into discrete high-density domains along z-axis. Over time, these domains gradually coalesce into larger slabs, exhibiting the characteristic morphological evolution of spinodal decomposition Cahn and Hilliard (1958).

To quantify the phase separation kinetics, we adopted a two-stage metric strategy. In the early stage, where interfaces are diffusive, we monitored the normalized density variance (⟨σ^ρ2⟩, see Supplementary Information) derived from the instantaneous density profile ρ(z) to track the growth rate of density fluctuations. As clear interfaces emerged, we switched to tracking the average slab size (⟨ξ⟩), defining dense domains as continuous regions where ρ(z) exceeds a threshold value (here the average system density was used as the threshold). To ensure statistical robustness, all kinetic metrics were averaged over six independent slab simulation replicas. The resulting kinetic profiles (Figure 4F and G) reveal two distinct regimes. First, immediately after the temperature quench, the system undergoes a rapid exponential growth in density variance (top panels). This behavior is characteristic of spinodal decomposition Cahn (1961). The results show that stronger effective attraction (higher ϵss or lower ϵb) tends to accelerate these initial fluctuations. This is quantitatively evidenced by the monotonic decrease in the characteristic time scale τ1/2, defined as the time required to reach 50% of the maximum normalized variance. Second, in the late stage (t > 10 ns), the domain growth transitions into a coarsening regime following a power-law scaling ⟨ξ⟩ ∼ tα with α ≈ 0.26 (Bottom panels). Theoretically, the Brownian motion coalescence (BMC) mechanism predicts an asymptotic scaling of α = 1/3 Majumdar et al. (1994). Our observed exponent falls slightly below this theoretical value, likely due to the viscoelastic nature of the polymer solution. Specifically, the sub-diffusive dynamics of individual chains (TAMSD ∼ Δt0.91) inherently constrain the macroscopic domain diffusion Doi et al. (1988), thereby retarding the coarsening rate below the standard Brownian prediction.

Intriguingly, between the initial spinodal decomposition and the late-stage Brownian coalescence, we observed a distinct intermediate plateau regime where the domain growth is temporarily arrested. As shown in the slab size evolution (Figure 4F and G, bottom), the domain growth stalls for a distinct period before entering the late-stage scaling regime. This kinetic arrest is visually corroborated by the density profiles at 5 ns in Figure 4E, where distinct domains are observed to form but fail to fuse immediately. The quantitative analysis in Figure 4F and G further reveals that these domains persist as distinct entities for durations that vary significantly depending on the desolvation parameters. This phenomenon is consistent with the characteristics of viscoelastic phase separation Tanaka (2000). It suggests that strong inter-chain interactions within the dense phase form a transient network structure. While surface tension drives the fusion of domains, the internal structural connectivity generates a resistance to the necessary shape deformation, thereby delaying coalescence. Crucially, the plateau duration extends significantly in systems with higher ϵss or lower ϵb. This trend corroborates our viscoelastic interpretation, suggesting that stronger inter-chain attractions reinforce the transient network, thereby effectively prolonging the kinetic arrest. This kinetic trap is particularly pronounced in the standard HPS model (red trajectories in Figure 4F and G), due to its relatively stronger pairwise attractions. The sensitivity of kinetic arrest of coarsening dynamics to desolvation parameters underscores the explicit modeling of desolvation in coarse-grained model in order to achieve a more physically plausible molecular simulations of LLPS. Synthesizing these observations, Figure 4E provides a schematic illustration of the three-stage evolution. This diagram maps the progression of driving forces, highlighting the transition from the initial regime dominated by inter-chain interactions, to a final stage governed primarily by surface tension-driven coarsening.

Collectively, we presented a comprehensive analysis of the phase separation dynamics, bridging the microscopic diffusion of single chains to the macroscopic growth of condensed domains. Our analysis underlies the essential role of desolvation potential. Beyond modulating thermodynamic stability and conformational ensembles as discussed earlier, it acts as a critical regulator of phase separation dynamics.

Parameterization of desolvation model for IDPs

The results from the homopolymer model discussed above underscore the critical role of desolvation in regulating both the thermodynamics and dynamics of LLPS. Consequently, explicitly incorporating water-mediated interactions into residue-level coarse-grained models is essential for more realistically capturing the physics of protein phase separation. To this end, we developed a transferable parameterization strategy derived from all-atom simulations and validated it against experimental observables of IDPs.

Experimentally determining the desolvation parameters, specifically, the barrier height ϵb and the solvent-separated minimum depth ϵss, remains challenging. To bridge this gap, we leveraged all-atom molecular dynamics simulations of amino acid analogues to extract the corresponding energy scales. Taking the depth of the direct-contact potential well (ϵ) as a reference scale, we expressed the desolvation parameters as ϵb = αbϵ and ϵss = αssϵ. By fitting the PMF profiles obtained from all-atom simulations (Figure 1—figure supplement 1), we determined the coefficients αb and αss. Notably, these coefficients were highly consistent across different pairs of amino acid analogues. For simplicity and transferability, we adopted the averaged values αb = 0.33 and αss = 0.06 as the initial baseline for our coarse-grained model.

To evaluate the effectiveness of these physics-derived parameters, we first integrated them into the framework of the standard HPS model. While the HPS model performs remarkably well in capturing sequence-dependent properties of LLPS, previous studies have noted that parameters optimized solely against single-chain Rg can face challenges in transferability to the condensed phase, often leading to an overestimation of phase separation propensity (i.e., higher critical temperatures, Tc) Dignon et al. (2018b). We assessed the model performance on a dataset of 47 IDPs taken from Tesei et al. Tesei et al. (2021). The deviation of simulated single-chain Rg from experimental SAXS data was quantified by χRg2=[(Rg,exp−Rg,sim)/σexp]2, where σexp represents the experimental uncertainty. When the desolvation parameters derived from all-atom simulations (αb = 0.33, αss = 0.06) were added to the HPS model without modifying its original ϵ, we observed a substantial improvement in conformational accuracy. The average discrepancy ⟨χRg2⟩ decreased drastically from 1,082.27 (original HPS) to 85.88 (HPS-desolvation) (Figure 5B). This improvement indicates that introducing the desolvation terms effectively mitigates the over-compaction tendency of the original HPS model likely by the competition between direct and water-separated contacts.

Parameterization of desolvation terms for the HPS model and CALVADOS2 model based on IDPs.

(A) Schematic workflow of the desolvation parameterization. (B) Correlation between experimental Rg and simulation Rg for the original HPS model (blue) and the revised HPS model with default desolvation scales (α = 0.33 and α = 0.06) (purple). The ⟨χRg2⟩ for different models were also shown. (C) Correlation between experimental Rg and simulation Rg for the original CALVADOS2 model (orange), the revised CALVADOS2 model with default desolvation (yellow), and revised CALVADOS2 model with optimized desolvation (αb = 0.3, αss = 0.03 and ϵ = 0.262 kcal/mol)(red). (D) Coexistence curves of FUS LC simulated with the original HPS model (blue), the revised HPS model with default desolvation scales (purple), the energy-rescaled HPS model (green), the default CALVADOS2 model (orange), and the revised CALVADOS2 model with optimized desolvation scales (red). The green shaded regions highlight the deviations of the desolvation models relative to the original frameworks.

We also incorporated the desolvation terms into CALVADOS2 Tesei and Lindorff-Larsen (2022), a force field whose hydrophobicity parameters (λ) were reparameterized using Bayesian optimization to reproduce experimental Rg and PRE data. Directly adding desolvation terms to CALVADOS2 led to substantial chain over-expansion (⟨χRg2⟩=643.30), reflecting the fact that CALVADOS2 is already finely tuned to reproduce experimental data through a top-down optimization procedure and the inclusion of additional desolvation terms is incompatible with its parameterization. To address this, we implemented a systematic optimization strategy in which the desolvation barrier height (ϵb), solvent-separated minimum depth (ϵss), and the overall energy factor ϵ were scanned over broad parameter ranges to identify the parameter set that minimizes deviations between simulated and experimental Rg values. The full optimization workflow is summarized in the flowchart in Figure 5A, and the resulting error landscapes are shown in Figure 5—figure supplement 1 and 2. The optimized parameter set (Table 1) achieved excellent agreement with experiments, reducing ⟨χRg2⟩ to 9.54 (Figure 5C), which is comparable to the accuracy of the original CALVADOS2 model in describing Rg values.

We further probed the phase behavior using the FUS low-complexity domain (FUS LC) as a test case. Experimental observations indicate that the critical temperature is around 25 ◦C Burke et al. (2015). While the original HPS model overestimates this value with a critical temperature Tc ∼ 340 K, integrating the desolvation terms into the HPS framework (HPS-desolvation) results in a Tc of ∼ 290 K (Figure 5D) which aligns well with experimental observations. In addition to improving the prediction of the critical temperature, we observed that the desolvation model significantly mitigates the overestimation of condensate density common in implicit solvent coarse-grained simulations Regy et al. (2021); Benayad et al. (2020). In the CALVADOS2 framework, the inclusion of desolvation reduces the dense phase density while maintaining a critical temperature comparable to the original model (Figure 5D). Similarly, for the HPS model, when we rescaled the interaction strength ϵ to match the single-chain Rg and thereby align the critical temperature with the HPS-desolvation system (Figure 5—figure supplement 3), we observed the same behavior. Quantitatively, across both frameworks, the desolvation-modulated models consistently yield a packing density approximately 5% lower than the original models. This distinction underscores a limitation of standard Lennard-Jones potentials where a single parameter ϵ simultaneously controls chain dimension, critical temperature, and condensate density. In contrast, our model introduces tunable parameters ϵb and ϵss that allow for accurate reproduction of experimental densities without compromising the prediction of phase boundaries. The observed density reduction is physically attributed to the increased population of solvent-mediated contacts relative to direct contacts. This mechanism introduces water-mediated porosity into the droplet and mimics the realistic hydration properties of protein condensates, thereby offering a more physically realistic description of IDP phase separation.

Discussions and Conclusion

Coarse-grained molecular simulations have become indispensable for elucidating the molecular principles underlying LLPS, complementing both theoretical frameworks and experimental observations. While residue-level models with implicit solvent have achieved remarkable success in capturing sequence-dependent phase behaviors of biomolecules, they often face a fundamental tradeoff due to the neglect of explicit solvent excluded volume and hydration effects. Specifically, interaction parameters calibrated to match single-chain dimensions or critical temperatures frequently result in an overestimation of the condensed phase density. This limitation suggests that standard Lennard-Jones potentials, which couple phase stability and density through a single energetic parameter, may lack the physical dimensionality required to fully describe the thermodynamics of hydrated protein condensates.

To address this challenge, we developed a coarse-grained framework that explicitly incorporates desolvation effects into the pairwise energy function. Application of this model to homopolymer systems demonstrates that desolvation substantially reshapes phase behavior. Incorporating desolvation modulates the critical temperature and alters the phase diagram by regulating intermolecular packing within the dense phase. Moreover, the model reveals a tight coupling between desolvation-controlled chain compaction during the formation of condensate phase and the thermal distance from criticality (temperature gap), providing a theoretical basis for linking single-chain conformational changes (ΔRg) to macroscopic phase behavior. Dynamic analysis further captures the three characteristic stages of the early stage of LLPS, including spinodal decomposition, kinetic arrest, and coarsening. Crucially, it reveals how desolvation effect regulates the kinetics of these stages by modulating the effective inter-chain interactions and the roughness of the energy landscape. These results underscore the importance of accounting for desolvation energetics in governing both the thermodynamics and kinetics of LLPS.

Motivated by these insights, we further developed a transferable parameterization strategy for residue-level models. The resulting desolvation-aware CG model improves the accuracy of IDP conformational ensembles and LLPS thermodynamics while preserving the computational efficiency inherent to CG methodologies. By bridging the gap between the computational efficiency of implicit-solvent representations and the physicochemical realism of water-mediated interactions, this framework provides a mechanistic means to decouple the control of phase stability (Tc) from the regulation of condensed-phase density (ρdense). This decoupling effectively mitigates the over-compaction of dense phase commonly observed in implicit solvent simulations with CG models, enabling the molecular simulations qualitatively more consistent with the physical nature of biological condensates.

Despite these advances, the current model employs a single set of desolvation parameters (αb, αss) for all amino acid types. While this simplification is effective for capturing global phase behavior, it overlooks residue-specific desolvation energetics that may influence selective interactions in heterogeneous or compositionally complex sequences. Future extensions of the model could incorporate residue-specific desolvation parameters derived from bottom-up parameterization or expanded experimental datasets, thereby enhancing predictive accuracy for sequence-dependent LLPS.

In summary, this work introduces a versatile and transferable extension to existing coarsegrained force fields that explicitly incorporates the physics of water-mediated interactions. By enabling efficient and realistic treatment of desolvation, the model provides a robust framework for simulating biomolecular condensates with improved thermodynamic and structural fidelity, as well as more accurate kinetic characterization. Future developments, such as residue-specific desolvation energetics and extensions to multicomponent systems involving nucleic acids, hold strong potential for deepening our mechanistic understanding of the complex phase behaviors that organize cellular biochemistry.

Methods

All-atom MD simulations

The all-atom MD simulations were performed using GROMACS. Ten amino acid analogues were solvated in a cubic water box under periodic boundary conditions. The OPLS-AA force field was used to model the solute molecules, and the TIP4P water model was employed to represent solvent. Na+ and Cl− ions were added to neutralize the system, yielding a total system size of 1,010 atoms. After an initial energy minimization of 5 × 104 steps, the system was equilibrated in the NVT ensemble at 298 K for 100 ps, followed by an NPT equilibration at 298 K and 1 atm for 1 ns. Temperature and pressure were controlled using the Nosé–Hoover thermostat and Parrinello–Rahman barostat Parrinello and Rahman (1981), respectively. A time step of 2 fs was used for all MD steps. The simulation details are provided in Table 2. Subsequently, a production simulation of 100 ns was performed under the same NPT conditions for structural and thermodynamic analysis. The intermolecular separation r between two amino acid analogues was defined as the distance between their C atoms. The potential of mean force (PMF) was computed as PMF(r) = −kBT ln P (r), where P (r) is the probability distribution of intermolecular distances obtained from the production trajectory.

Coarse-grained simulations

We utilized the slab method Dignon et al. (2018b) to extract the densities of the coexisting dilute and dense phases. In these simulations, 100 protein chains were placed in a rectangular box of dimensions 15 nm × 15 nm × 280 nm. The density profile along the z-axis was computed to identify the condensed and dilute regions, from which thermodynamic quantities such as the critical temperature were subsequently determined. Simulations were performed over a range of different temperatures. To implement the pairwise interaction potentials of the HPS model, we employed the azplugins package (https://github.com/mphowardlab/azplugins) together with HOOMD-blue v2.9.7 Anderson et al. (2020), running with GPU acceleration. All simulations used a time step of 0.01 ps. Each trajectory was first equilibrated for 3 × 107 steps, followed by a production run of 1 × 108 steps. Additional details of the simulation framework are provided in Table 3.

Hydrophobicity Scale Model

Our desolvation model is developed based on the hydrophobicity scale (HPS) coarse-grained model by Dignon et al. (2018b), in which each residue is represented by a single bead. The potential energy includes bonded term, electrostatic term and short-range pairwise term. The bonded interaction term is given by a harmonic potential V b = kb(ri,i+1 − r0)2 with the spring constant kb = 10 kJ/Å2 and the bonded length r0 = 0.38 Å. ri,i+1 represents the distance between the consecutive residues i and i + 1 along the protein chain. The electrostatic interaction term is modeled by a Coulombic potential with Debye-Hückel Debye and Hückel (1923) electrostatic screening caused by salt ion concentration. The function is given by (3) Vijele(r)=qiqj4πDrexp⁡(−r/κ), where κ is the screening length and D = 80, which is the relative dielectric constant of water. For the short-range pairwise interaction, the HPS model uses a 6-12 Lennard-Jones potential with the interaction strength given by hydrophobicity scale Kapcha and Rossky (2014), which can describe the effective interaction between residues. The hydrophobicity value λ, ranging from 0 to 1, varies for different residues. The arithmetic average is used for the interaction parameter between two residues, i.e. hydrophobicity value λi,j = (λi + λj)/2 and amino acid size σi,j = (σi + σj)/2. The energy function of the short-range pairwise interaction is given by (4) Φ(r)={ΦLJ+(1−λ)ϵ,ifr⩽21/6σλΦLJ,otherwise where ΦLJ is standard 6-12 Lennard-Jones potential (5) ΦLJ=4ϵ[(σr)12−(σr)6].

The free parameter ϵ is set as 0.2 kcal/mol, with which the molecular simulations can best reproduce the experimental results of Rg values for a set of IDPs.

Simulation Framework

We use slab simulation scheme in the coexistence simulations. Initially, 100 chains are randomly distributed in a cubic box with periodic boundary condition at 200 K. The box length is first scaled to 15 nm, which can effectively prevent protein chains from interacting with their periodic images generated by periodic boundary conditions. Then, the x- and y-dimensions are set fixed and the z-dimension is extended to 280 nm. The temperature is then gradually increased to the target temperature and the simulations are conducted at different temperatures for 1 μs in the NVT ensemble maintained by a Langevin thermostat, with a friction coefficient of 0.01 ps−1. A time step of 0.01 ps is used for all the simulations. We use HOOMD-Blue v2.9.7 packages Anderson et al. (2020) to conduct the simulations, which utilizes both CPU and GPU resources to accelerate calculation.

The slab density profile along z-axis is then determined after the system gets equilibrated. In order to avoid the influence of periodic boundary conditions, we first move the center of mass of the system in the z-axis direction to the position of z = 0. We then determined the distribution of dense phase and dilute phase of the phase separated system using the defined thresholds. The area with density higher than 0.95 of the maximum density (ρmax) is treated as dense phase, and their average density is regarded as the phase density of the dense phase (ρdense). For the dilute phase, the threshold is taken as ρmin + 50 mg/mL, where ρmin is the minimum density. The density of dilute phase (ρdilute) is then determined by the same method. If ρmax − ρmin < 50 mg/mL, the system is considered not to undergo phase separation.

Critical Temperature

The critical temperature was determined based on the vapor-liquid interfacial properties of Lennard-Jones chains Blas et al. (2008). The critical temperature Tc is obtained from ρdense and ρdilute at different temperatures by (6) ρdense−ρdilute=A(Tc−T)β, where β is the critical exponent, which is set as 0.325 Rowlinson and Widom (2013) and A is the fitting parameter. Data with temperature below the critical temperature is used to fit this equation. The phase density at critical temperature is determined by the law of rectilinear diameters Davies (1912) (7) ρdense+ρdilute2=B+C⋅T, with B and C being the free parameters determined from the simulation data. At the critical temperature, the distinction between dense phase and dilute phase disappears, which means ρdense = ρdilute = ρc. The phase density is then calculated by ρc = B + C ⋅ Tc. A phase diagram which shows temperature change with the density of dense phase and dilute phase is then obtained.

Derivation of the Reduced Diffusion Coefficient

To investigate the intrinsic influence of the interaction potential on chain dynamics, we must isolate the structural contributions from the trivial thermal acceleration. While keeping the quench depth (T/Tc) constant ensures thermodynamic consistency, it necessitates varying the absolute temperature T, which inherently alters the base diffusion rate via thermal fluctuations. To resolve this, we adopt the constant quench depth approach to ensure thermodynamic consistency and introduce a normalized metric to eliminate the trivial thermal velocity contribution.

According to the fundamental Einstein relation, which connects the diffusion coefficient K to the thermal energy kBT and the friction coefficient ζ: (8) K(T)=kBTζ(T,ϕ,ϵ)

Here, the friction coefficient ζ encapsulates all resistance to motion. Following the hybrid theoretical framework proposed by Macedo and Litovitz Macedo and Litovitz (1965), which unifies free-volume theory with activated rate theory, the friction term can be factorized into a packing-dependent contribution and an energy-barrier-dependent contribution: (9) ζ(T,ϕ,ϵ)∝exp⁡(γv∗vf(ϕ))×exp⁡(Eeff(ϵ)kBT) where the first term represents the hindrance due to limited free volume (vf), governed by the packing density ϕ, and the second term is the Arrhenius factor describing the activated hopping over an effective energy barrier Eeff, which corresponds to the roughness of the energy landscape induced by the desolvation potential.

In our analysis, by fixing the quench depth (T/Tc), we ensure that the macroscopic packing density ϕ remains comparable across different systems (as verified by the phase diagram in main text). Consequently, the free-volume contribution to friction is effectively invariant. Substituting this back into Equation (8), the diffusion coefficient scales as: (10) K(T)∼T⋅exp⁡(−Eeff(ϵ)kBT)

The pre-factor T represents the trivial thermal acceleration inherent to Brownian motion. To isolate the intrinsic energy landscape signature, we define the reduced diffusion coefficient K~ by normalizing out this linear thermal contribution: (11) K~≡KT∝exp⁡(−Eeff(ϵ)kBT)

Thus,K~ serves as a physically rigorous metric to isolate the intrinsic influence of the energy function on chain dynamics. By normalizing out the trivial linear thermal velocity contribution inherent to Brownian motion, this definition is necessary to decouple the specific kinetic modulation imposed by the desolvation potential from the background thermal acceleration and thermodynamic driving forces. Consequently, K~ allows for a direct assessment of how the interaction potential fundamentally alters transport properties, independent of the trivial temperature effects required to maintain thermodynamic consistency.

Calculation of Normalized Density Variance

To quantify the rate of density fluctuation growth during the early stage of phase separation, we first computed the instantaneous spatial variance of the density profile along the z-axis, σρ2(t), defined as (12) σρ2(t)=1Lz∫0Lz[ρ(z,t)−ρ¯]2dz where ρ(z, t) is the local density at position z and time t, and ρ¯ is the mean density of the system. To enable a fair comparison of kinetics across systems with distinct interaction parameters (ϵss and ϵb), it is crucial to decouple the rate of phase separation from the difference in equilibrium densephase densities. Therefore, we normalized the instantaneous variance by its equilibrium baseline: (13) σ^ρ2(t)=σρ2(t)⟨σρ2⟩eq.

Here, the normalization factor ⟨σρ2⟩eq is the time-averaged density variance calculated from the final 30% of the simulation process. This period corresponds to the steady state characterized by the formation of a well-defined slab with stabilized phase densities. By using this normalized metric σ^ρ2(t), we effectively factor out the variations in final density contrast, ensuring that the observed differences in the time-evolution curves purely reflect the acceleration or deceleration of the kinetic process itself.

Desolvation Parameters

Due to the lack of experimental data, the empirical parameters αb and αss were estimated using all-atom molecular dynamics simulations of amino acid analogues with explicit solvent. The potential of mean force (PMF) obtained from the simulation results is shown in Figure 1—figure supplement 1. Methane (CH4), methanol (CH3OH), acetate ion (C2H3O2−), ammonium ion (NH4+), and methanamide (CH3NO) were chosen as representative amino acid analogues for non-polar, polar, negatively charged, positively charged, and backbone-like residues, respectively. The αb, αss values were then obtained by fitting the nonbonded energy function with desolvation effect (Equation (1)) to the experimental PMF results. The λ parameter values for these amino acid analogous were assigned based on the λ values of the corresponding amino acids. Specifically, we used λGLY = 0.649 for methane, λSER = 0.595 for methanol and λCYS = 0.595 for methanamide. The results are shown in Table 1.

eLife Assessment

This manuscript presents a valuable and timely contribution by incorporating desolvation barriers into coarse-grained models of biomolecular condensates. The findings are convincing, supported by a clear physical model and systematic simulations showing effects on phase behavior, packing, and dynamics. Some clarification and broader context would improve the manuscript, but it provides a foundation that will be of use for developing more realistic coarse-grained interaction schemes.

Reviewer #1 (Public review):

This manuscript is very interesting and timely. By introducing the critical effects of desolvation barriers and solvent (water)-separated minima into the implicit-solvent potentials (of mean force, PMFs) for coarse-grained molecular dynamics simulations of biomolecular liquid-liquid phase separation (LLPS), this work fills a gap that should be apparent to researchers of protein folding in the past couple of decades but has so far escaped deserved attention such that these basic features of aqueous solvation have seldom, though not never, been invoked in recent studies of biomolecular condensates. Although the present paper deals almost exclusively with homopolymers, this work can be a foundation for the future development of a new, more physical coarse-grained interaction scheme for simulating amino acid sequence-dependent effects, which I presume is the authors' ongoing or next endeavor. The results presented in this manuscript are highly valuable.

However, there is room for improvement in the authors' description of (i) the broader impact of effects of desolvation barrier and solvent-separated minimum in the thermodynamics of biomolecular condensates, especially with regard to the ramifications on hydrostatic pressure-dependent effects; (ii) the physical implication of using a 20-parameter hydropathy scale rather than a 210-parameter pairwise amino acid interaction scheme; and (iii) temperature-dependent effects, including the authors' discussion of "enthalpic" and "entropic" contributions. In all these aspects, the authors' discussion should be put in a more comprehensive context of the existing literature. At a few other places, the description of the methods and results should be clarified as well. Accordingly, the authors should revise the manuscript to address the following items thoroughly within the revised manuscript (not merely in the response letter) with the additional references mentioned below included in the revised discussion:

(1) In several places, e.g., on line 77 (p.2), the authors appear to suggest that "implicit-solvent representation" is the origin of the deficiency in commonly utilized coarse-grained potentials that this study is aiming to rectify. But desolvation barriers and solvent-separated minima are also features of implicit-solvent representations; they are just features that should be incorporated in more accurate implicit-solvent potentials. This point is stated quite clearly and accurately in the Abstract (p.1) but not consistently in the rest of the text. The authors should check the entire text carefully to ensure that a coherent, accurate perspective is presented.

(2) In the discussion of the importance of desolvation barriers and solvent-separated minima in the Introduction (pp.1-3), connections should be drawn to recent works that utilize these PMF features to rationalize hydrostatic pressure (P)-modulated effects on biomolecular LLPS, including the P-dependent reentrant phase separation of alpha elastin; see Cinar et al. (2019) Chem Eur J 25:13049 (https://chemistry-europe.onlinelibrary.wiley.com/doi/full/10.1002/chem.201902210) and references therein, especially discussions around Figures 10, 11 & 13 in this reference.

(3) In the lower panels of Figures 2D, E (p.5), what do the differently colored small circles in the double-minimum free energy profiles represent? Does the color shading have the same meaning as that in the upper panels? If so, what do the positions of the circles on the free energy profile represent? The authors should clarify this.

(4) The discussion regarding entropy and enthalpy around Figure 2 is quite confusing as it stands. What do the authors mean exactly by the association of entropy or enthalpy with the desolvation barrier of the solvent-separated minimum? Are they referring to conformational entropy?

(5) Do the authors assume that the PMF (effective implicit-solvent potential) is a purely enthalpic term? It appears to be the authors' assumption. If so, the assumption has to be stated clearly in their discussion of "entropy" vs "enthalpy" around Figure 2.

(6) Closely related to points 3-5 above, it should be stated clearly that the "temperature" used in the authors' simulations does not represent experimental temperature if the authors are using purely enthalpic effective potentials because PMFs are in fact temperature-dependent. This clarification is necessary to avoid misunderstanding. In this regard, it should be noted that temperature-dependent effective interactions have been used for modeling biomolecular condensates in analytical theory (Lin, Song, Forman-Kay & Chan, J Mol Liq 2017, already in the citation list) as well as in coarse-grained molecular dynamics simulations [Dignon et al. (2019) ACS Cent Sci 5:821-830 (https://pubs.acs.org/doi/10.1021/acscentsci.9b00102); Chakravarti & Joseph (2025) Protein Sci 34:e70284 (https://onlinelibrary.wiley.com/doi/10.1002/pro.70284)]. The latter two studies, not cited currently, are particularly relevant and thus should be cited because the authors may wish to incorporate temperature-dependent features in their ongoing or future effort in constructing a more comprehensive coarse-grained interaction scheme for biomolecular LLPS simulation.

(7) In tackling "entropy" vs "enthalpy", it should be noted that the temperature dependence of the effective interactions entails an entropic contribution (which is itself temperature dependent) in addition to conformational entropy. As for the effective potential with desolvation barrier and solvent-separated minimum, it should be noted that the decomposition into entropic and enthalpic contributions at the direct contact, desolvation barrier, and solvent-separated minimum can be dramatically different, see, e.g., MaCallum et al. (2007) PNAS 104:6206-6210 (https://www.pnas.org/doi/full/10.1073/pnas.0605859104) and references therein.

(8) P.7, line 340: The proportionality relation follows directly from the standard Flory-Huggins result T_c = T chi(T)/chi_c, thus the proportionality constant is exactly 1/chi_c. Is this the standard relation that the authors are invoking here? The authors should clarify this.

(9) The study on dynamic consequences on pp.8-11 is interesting, but clarifications are necessary:

(i) The vertical schematic in Figure 4A should be explained in detail in its entirety. As it stands, no explanation is provided either in the figure caption or in the text. In particular, what does "elasticity driven" refer to?

(ii) The top snapshot in Figure 4A is labeled t_sim = 0 ns. Does it mean that the snapshot shown is the only chain configuration that the authors used to start the simulation, and that the snapshot does NOT represent the result of any time evolution, no matter how short the duration is? However, if that is the case, why is this snapshot identified with spinodal decomposition if it is not the product of a time evolution from a more homogeneous configuration?

(iii) Related to (ii) - do the rectangular boxes shown represent the entire simulation box or just part of the box containing the polymer chains? One would imagine that if the top snapshot represents spinodal decomposition, the simulation would have been started at a more uniform distribution a short time prior? Why is this not the case?

(iv) What precisely do the small yellow beads and black-colored springs in the zoom-in image of Figure 4E represent?

(10) In discussing dynamic effects, it is useful to draw connections to related works on the effect of chain flexibility on "aging" of condensate [Biswas & Potoyan (2024) PRX 45:9222-9245 (https://journals.aps.org/prxlife/abstract/10.1103/PRXLife.2.023011)] and characterization of viscoelasticity in simulations of biomolecular condensates [Tejedor et al. (2023) J Phys Chem B 127:4441-4459 (https://pubs.acs.org/doi/10.1021/acs.jpcb.3c01292)], as the effects of desolvation can be explored further based on these prior works.

(11) Much of the present study is based on the original HPS formulation of Dignon et al. (2018). In this regard and also in anticipation of future development of improved interaction schemes, several issues should be stated and discussed, even if briefly:

(i) The original HPS model has a basic shortcoming in accounting for the relative interaction strengths of, among others, arginine vs lysine residues [Das et al. (2020) PNAS 117:28795-28805 (https://www.pnas.org/doi/10.1073/pnas.2008122117)].

(ii) Compared to 210-parameter pairwise interaction schemes, such as KH in Dignon et al. (2018) and Joseph et al. (2021), the 20-parameter interaction scheme is likely too restrictive to account for pairwise amino acid residue interactions [Wessén et al. (2022) J Phys Chem B 45:9222-9245 (https://pubs.acs.org/doi/10.1021/acs.jpcb.2c06181)].

(iii) The height of the desolvation barrier may vary significantly for different amino acid residue pairs, see, e.g., Figure 11 of Cinar et al. (2019) mentioned above (and references therein). The authors should discuss these nuances in the revised version. They may also wish to take them into consideration in future investigations.

Reviewer #2 (Public review):

Summary:

This manuscript addresses an important and timely question in the molecular simulation of biomolecular condensates. Most residue-level coarse-grained models used for IDP phase separation employ implicit solvent and represent effective interactions through relatively simple pairwise potentials. While these models have been very useful, they usually do not explicitly distinguish direct contacts from solvent-separated interactions, nor do they include an energetic barrier associated with water removal. This manuscript attempts to address that limitation by introducing desolvation-inspired terms into coarse-grained models and examining their consequences for phase behavior, chain conformations, dense-phase packing, and dynamics.

Strengths:

The central idea is physically well motivated. Using a simple homopolymer model, the authors show that increasing the desolvation barrier suppresses phase separation, whereas stabilizing solvent-separated contacts enhances phase separation. They further show that solvent-separated interactions can reduce dense-phase over-compaction, which is a meaningful result given the known challenges in obtaining both accurate single-chain dimensions and realistic dense-phase properties from the same coarse-grained model. The finding that desolvation-like terms can reshape dense-phase packing without simply rescaling the overall interaction strength is interesting and could be useful for future model development. I also found the attempt to connect conformational changes across dilute and dense phases with thermal distance from the critical point to be intriguing. The dynamic analysis, including the FRAP-like simulations and the discussion of kinetic arrest during coarsening, adds another useful dimension to the work.

Weaknesses:

At the same time, there are several places where the manuscript would benefit from more careful framing. First, the desolvation terms are still effective coarse-grained parameters rather than a direct representation of water molecules. The language sometimes gives the impression that desolvation is being treated explicitly, whereas the model introduces desolvation-inspired effective interactions into an implicit-solvent framework. Second, the conformational analysis is interesting, but the broader context of prior work on dilute-to-dense phase conformational reorganization of IDPs could be more clearly discussed. This would help clarify what is new in the present work, whether it is the conformational change itself, its dependence on desolvation terms, or the proposed scaling with distance from the critical point. Third, the dynamic results are potentially useful, but the manuscript should more clearly articulate what is nontrivial beyond the expected slowing of local rearrangements by an added barrier in the potential.

Overall, I think this is a useful and potentially important contribution.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

This manuscript is very interesting and timely. By introducing the critical effects of desolvation barriers and solvent (water)-separated minima into the implicit-solvent potentials (of mean force, PMFs) for coarse-grained molecular dynamics simulations of biomolecular liquid-liquid phase separation (LLPS), this work fills a gap that should be apparent to researchers of protein folding in the past couple of decades but has so far escaped deserved attention such that these basic features of aqueous solvation have seldom, though not never, been invoked in recent studies of biomolecular condensates. Although the present paper deals almost exclusively with homopolymers, this work can be a foundation for the future development of a new, more physical coarse-grained interaction scheme for simulating amino acid sequence-dependent effects, which I presume is the authors' ongoing or next endeavor. The results presented in this manuscript are highly valuable.

We thank the reviewer for all these positive comments.

However, there is room for improvement in the authors' description of (i) the broader impact of effects of desolvation barrier and solvent-separated minimum in the thermodynamics of biomolecular condensates, especially with regard to the ramifications on hydrostatic pressure-dependent effects; (ii) the physical implication of using a 20-parameter hydropathy scale rather than a 210-parameter pairwise amino acid interaction scheme; and (iii) temperature-dependent effects, including the authors' discussion of "enthalpic" and "entropic" contributions. In all these aspects, the authors' discussion should be put in a more comprehensive context of the existing literature. At a few other places, the description of the methods and results should be clarified as well. Accordingly, the authors should revise the manuscript to address the following items thoroughly within the revised manuscript (not merely in the response letter) with the additional references mentioned below included in the revised discussion:

(1) In several places, e.g., on line 77 (p.2), the authors appear to suggest that "implicit-solvent representation" is the origin of the deficiency in commonly utilized coarse-grained potentials that this study is aiming to rectify. But desolvation barriers and solvent-separated minima are also features of implicit-solvent representations; they are just features that should be incorporated in more accurate implicit-solvent potentials. This point is stated quite clearly and accurately in the Abstract (p.1) but not consistently in the rest of the text. The authors should check the entire text carefully to ensure that a coherent, accurate perspective is presented.

We thank the reviewer for pointing out this important issue. We agree that implicit-solvent representation itself is not the origin of the deficiency. Our intention is to incorporate desolvation-inspired effective terms within an implicit-solvent coarse-grained framework, because many commonly used implicit-solvent potentials do not directly account for the desolvation features, such as the desolvation barrier and the solvent-separated potential well.

We have revised the Abstract, Introduction, Results, and Discussion to make this distinction consistent throughout the manuscript. The revised text now emphasizes that the model remains an implicit-solvent CG model, but contains additional effective terms inspired by desolvation features observed in all-atom PMFs.

Corresponding changes:

(1) (page 1, lines 17–20) The Abstract identifies the model as an implicit-solvent CG model with added desolvation terms:

"Here, guided by all-atom simulations and experimental measurements, we develop a desolvation-aware implicit-solvent CG model by incorporating residue-level desolvation terms directly into the pairwise energy function and apply it to investigate LLPS of intrinsically disordered proteins."

(2) (page 2, lines 80–81) The Introduction retains the implicit-solvent description of existing residue-level CG models:

"Despite these advances, most residue-level CG models rely on implicit solvent representations, in which individual water molecules are not explicitly represented."

(3) (page 2, lines 87–89) The specific limitation is identified as the absence of a direct account of the multi-step desolvation process:

"More importantly, conventional implicit-solvent CG models used for LLPS usually do not directly account for the multi-step desolvation process that accompanies the transition from a dilute solution to a dense condensate."

(4) (page 4, lines 172–174) The added terms are described as part of a desolvation-inspired effective potential:

"Together, these parameters shape the desolvation-inspired effective potential and modulate the statistical balance between direct-contact and solvent-separated configurations."

(5) (page 15, lines 521–522) The Discussion restates the implicit-solvent model framework:

"To address this challenge, we developed a desolvation-aware implicit-solvent CG framework that incorporates desolvation barrier and solvent-separated terms into the pairwise potential."

(2) In the discussion of the importance of desolvation barriers and solvent-separated minima in the Introduction (pp.1-3), connections should be drawn to recent works that utilize these PMF features to rationalize hydrostatic pressure (P)-modulated effects on biomolecular LLPS, including the P-dependent reentrant phase separation of alpha elastin; see Cinar et al. (2019) Chem Eur J 25:13049 (https://chemistryeurope.onlinelibrary.wiley.com/doi/full/10.1002/chem.201902210) and references therein, especially discussions around Figures 10, 11 & 13 in this reference.

We thank the reviewer for bringing this literature to our attention. We agree that pressure-modulated LLPS provides important context for the physical relevance of desolvation barriers and solvent-separated minima. We have therefore expanded the Introduction and Discussion to connect our model to prior work on hydrostaticpressure-dependent condensate behavior, including pressure-dependent reentrant phase separation of alpha-elastin.

Corresponding changes:

(1) (page 3, lines 93–95) The Introduction connects the PMF features to hydrostaticpressure-dependent LLPS:

"Related studies on hydrostatic pressure effects have further suggested that desolvation barriers and solvent-separated minima can help rationalise pressure-modulated LLPS behaviors, including the pressure-dependent reentrant phase separation of α-elastin Cinar et al. (2019, 2018)."

(2) (page 15, lines 533–537) The Discussion states the pressure-dependent implication conservatively:

"These findings may also provide a useful physical basis for future studies of pressure-dependent condensate behavior, as pressure-induced changes in hydration, solvent-separated states, and desolvation barriers have been proposed to contribute to pressure-modulated and reentrant LLPS Dias and Chan, 2014); Cinar et al. (2019, 2018)."

(3) In the lower panels of Figures 2D, E (p.5), what do the differently colored small circles in the double-minimum free energy profiles represent? Does the color shading have the same meaning as that in the upper panels? If so, what do the positions of the circles on the free energy profile represent? The authors should clarify this.

We thank the reviewer for identifying this ambiguity. The small circles in the lower panels of Figures 2D and 2E are qualitative schematic representations of residue-pair configurations along the effective pair-potential profile. Their blue and green colors distinguish the barrier-variation and solvent-separated-well cases, respectively; they are not a quantitative scale and do not encode temperature or population magnitude. The positions of the circles indicate the direct-contact, barrierregion, or solvent-separated regions, while the density of circles schematically represents the population of configurations.

We have clarified this interpretation in the Figure 2 caption and aligned the Results text with the redistribution among direct-contact, barrier-region, and solvent-separated states.

Corresponding changes:

(1) (page 6, Figure 2D and E lower panels) The schematics distinguish low and high ε_b or ε_ss and use the density and position of the circles to depict populations in the direct-contact, barrier-region, and solvent-separated regions; the blue and green colors distinguish the two parameter families and are not a quantitative scale.

(2) (page 6, Figure 2 caption) The caption defines the population encoding used in the lower panels:

"The lower panels schematically illustrate how changes in εb and εss alter the distribution of residue-pair configurations. The small circles indicate schematic populations of residue-pair configurations along the potential profile, with denser circles representing a higher population."

(3) (page 5, lines 210–213) The Results text connects the schematics to redistribution among the three residue-pair states:

"These opposing effects suggest that the desolvation potential regulates macroscopic phase behavior by redistributing residue-pair configurations between direct-contact, barrier-region, and solvent-separated states (lower panels of Figure 2D and E)."

(4) The discussion regarding entropy and enthalpy around Figure 2 is quite confusing as it stands. What do the authors mean exactly by the association of entropy or enthalpy with the desolvation barrier of the solvent-separated minimum? Are they referring to conformational entropy?

We thank the reviewer for pointing out this ambiguity. We agree that our original wording around entropy and enthalpy could be misleading, because it might imply a rigorous thermodynamic decomposition of the PMF. In the revised manuscript, we have therefore clarified that the effect of the desolvation barrier refers to an entropyrelated configurational restriction of residue-pair configurations near the barrier region, rather than the overall conformational entropy of the entire chain. We also replaced the previous "enthalpic stabilization" wording with "effective free-energy stabilization" to avoid implying that the solvent-separated minimum is treated as a purely enthalpic contribution.

Corresponding changes:

(1) (page 5, lines 199–201) The barrier effect is described in terms of the sampled residue-pair population:

"Analysis of residue-residue radial distribution functions showed that higher εb suppresses the population of configurations near the barrier region (Figure 2—figure Supplement 1C)."

(2) (page 5, lines 201–202) The entropy-related statement is restricted to configurational sampling near the barrier:

"This reduction in the statistical weight of barrier-region configurations can be interpreted as an entropy-related configurational restriction and thus disfavors phase separation."

(3) (page 5, lines 209–210) The solvent-separated minimum is described as an effective free-energy contribution:

"This solvent-separated minimum provides effective free-energy stabilization for water-mediated configurations and thereby promotes phase separation."

(4) (page 6, Figure 2D and E lower panels) The schematic headings are "Barrier-mediated Restriction" and "Solvent-separated Stabilization", avoiding a strict entropy-enthalpy decomposition.

(5) Do the authors assume that the PMF (effective implicit-solvent potential) is a purely enthalpic term? It appears to be the authors' assumption. If so, the assumption has to be stated clearly in their discussion of "entropy" vs "enthalpy" around Figure 2.

We thank the reviewer for raising this important point. We do not assume that the PMF obtained from all-atom simulations is a purely enthalpic term. The PMF is a free-energy profile that contains enthalpic and entropic contributions. The current manuscript defines the PMF as −kB T lnP(r), uses the all-atom PMFs to motivate a nonbonded effective coarse-grained potential, and describes the solvent-separated minimum as providing effective free-energy stabilization. We do not perform a rigorous enthalpy-entropy decomposition, and the revised wording avoids implying such a decomposition.

Corresponding changes:

(1) (page 16, lines 594–595) The Methods define the PMF as a free-energy profile obtained from the radial probability density:

"The potential of mean force (PMF) was computed as PMF(r) = −kBT ln P(r), where P(r) is the radial probability density obtained from the production trajectory."

(2) (page 5, Figure 1 caption) The CG interaction is labeled as an effective potential rather than as an enthalpic PMF decomposition:

"Pairwise effective potential incorporating desolvation-inspired terms. Different curves correspond to different desolvation parameters."

(3) (page 4, lines 172–174) The parameters are described as shaping an effective potential:

"Together, these parameters shape the desolvation-inspired effective potential and modulate the statistical balance between direct-contact and solvent-separated configurations."

(4) (page 5, lines 209–210) The solvent-separated contribution is described using free-energy language:

"This solvent-separated minimum provides effective free-energy stabilization for water-mediated configurations and thereby promotes phase separation."

(6) Closely related to points 3-5 above, it should be stated clearly that the "temperature" used in the authors' simulations does not represent experimental temperature if the authors are using purely enthalpic effective potentials because PMFs are in fact temperature-dependent. This clarification is necessary to avoid misunderstanding. In this regard, it should be noted that temperature-dependent effective interactions have been used for modeling biomolecular condensates in analytical theory (Lin, Song, Forman-Kay & Chan, J Mol Liq 2017, already in the citation list) as well as in coarse-grained molecular dynamics simulations [Dignon et al. (2019) ACS Cent Sci 5:821-830 (https://pubs.acs.org/doi/10.1021/acscentsci.9b00102); Chakravarti & Joseph (2025) Protein Sci 34:e70284 (https://onlinelibrary.wiley.com/doi/10.1002/pro.70284)]. The latter two studies, not cited currently, are particularly relevant and thus should be cited because the authors may wish to incorporate temperature-dependent features in their ongoing or future effort in constructing a more comprehensive coarse-grained interaction scheme for biomolecular LLPS simulation.

We agree with the reviewer that the simulation temperature should be interpreted carefully. In the present simulations, the effective potential is temperature-independent within each chosen parameter set. Therefore, the reduced temperature primarily serves as a model temperature controlling the relative strength of thermal fluctuations, rather than as a direct experimental temperature. We have clarified this point in the revised manuscript and added relevant references on temperature-dependent effective interactions, which represent an important direction for future model development.

Corresponding changes:

(1) (page 4, lines 177–179) The manuscript states that the absolute simulation temperature is not an experimental temperature:

"Because the effective interaction parameters used in the model are temperature-independent, the absolute simulation temperature should not be directly interpreted as an experimental temperature."

(2) (page 4, lines 182–184) The reduced temperature is identified as a model temperature:

"Accordingly, T* should be interpreted primarily as a model temperature that controls the relative strength of thermal fluctuations, rather than as having a direct quantitative correspondence with experimental temperature."

(3) (page 14, lines 481–485) The FUS LC temperature comparison is framed cautiously:

"It is worth noting that the residue-level coarse-grained models used here employ temperature-independent effective interaction parameters. As a result, the temperature values reported here cannot be interpreted as quantitatively equivalent to experimental temperatures, particularly when they deviate substantially from room-temperature conditions."

(4) (page 15, lines 563–567; continues on page 16, lines 568–569) The Discussion identifies temperature-dependent effective interactions and a corresponding future extension:

"In addition, effective interactions themselves can be temperature-dependent, as demonstrated in analytical theories and coarse-grained simulations of biomolecular condensates Lin et al. (2017); Dignon et al. (2019); Chakravarti and Joseph (2025). Future extensions of the model could therefore incorporate residue-specific and temperature-dependent desolvation parameters derived from bottom-up parameterization or expanded experimental datasets, thereby enhancing predictive accuracy for sequence-dependent LLPS."

(7) In tackling "entropy" vs "enthalpy", it should be noted that the temperature dependence of the effective interactions entails an entropic contribution (which is itself temperature dependent) in addition to conformational entropy. As for the effective potential with desolvation barrier and solvent-separated minimum, it should be noted that the decomposition into entropic and enthalpic contributions at the direct contact, desolvation barrier, and solvent-separated minimum can be dramatically different, see, e.g., MaCallum et al. (2007) PNAS 104:6206-6210 (https://www.pnas.org/doi/full/10.1073/pnas.0605859104) and references therein.

We thank the reviewer for this important clarification. We agree that temperature-dependent effective interactions can contain entropic contributions beyond conformational entropy and that the balance of enthalpic and entropic contributions may differ among the direct-contact minimum, desolvation barrier, and solvent-separated minimum. The present model does not decompose the PMF into temperature-dependent enthalpic and entropic components; accordingly, we have avoided assigning those components to individual PMF features. The Discussion cites explicit-solvent PMF analyses when noting residue-pair and temperature dependence and identifies temperature-dependent desolvation parameters as an important future extension.

Corresponding changes:

(1) (page 15, lines 561–563) The Discussion cites explicit-solvent PMF work when noting residue-pair and temperature dependence:

"Explicit-solvent PMF analyses have shown that desolvation barrier heights and solvent-separated minima can differ substantially among residue pairs and may also exhibit temperature dependence Cinar et al. (2019); Debiec et al. (2014); MacCallum et al. (2007)."

(2) (page 15, lines 563–566) The manuscript states that effective interactions can themselves depend on temperature:

"In addition, effective interactions themselves can be temperature dependent, as demonstrated in analytical theories and coarse-grained simulations of biomolecular condensates Lin et al. (2017); Dignon et al. (2019); Chakravarti and Joseph (2025)."

(3) (page 15, lines 566–567; continues on page 16, lines 568–569) Temperature-dependent desolvation parameters are identified as a future model extension:

"Future extensions of the model could therefore incorporate residue-specific and temperature-dependent desolvation parameters derived from bottom-up parameterization or expanded experimental datasets, thereby enhancing predictive accuracy for sequence-dependent LLPS."

(8) P.7, line 340: The proportionality relation follows directly from the standard FloryHuggins result T_c = T chi(T)/chi_c, thus the proportionality constant is exactly 1/chi_c. Is this the standard relation that the authors are invoking here? The authors should clarify this.

We thank the reviewer for pointing out the missing intermediate steps. Yes, the relation we invoked is based on the standard Flory-Huggins critical condition. In the revised manuscript, we have expanded the derivation to explicitly show how the critical condition chi(T_c) = chi_c leads to the relation between chi(T_sim) - chi_c and the normalized thermal distance (T_c - T_sim)/T_sim.

We also revised the wording to avoid presenting this as a universal law. The relation is now presented as a simulation-supported trend within the present model, rationalized by a simplified linear-response assumption between Delta R_g and the excess interaction strength.

Corresponding changes:

(1) (page 7, lines 263–264) The critical-condition substitution is now shown explicitly:

"At the critical point, χ(Tc) = χc, which gives εeff = kBTcχc. Substituting this relation into the expression for χ(Tsim) yields χ(Tsim) = χcTc/Tsim."

(2) (page 7, line 265) The resulting relation is written as Equation (2):

"χ(Tsim) − χc = χc (Tc − Tsim)/Tsim."

(3) (page 7, lines 266–267) The fixed-chain-length assumption is stated explicitly:

"For systems with the same chain length, χc is a fixed constant. Thus, the deviation from the critical interaction parameter is directly related to the rescaled thermal distance (Tc − Tsim)/Tsim."

(9) The study on dynamic consequences on pp.8-11 is interesting, but clarifications are necessary:

(i) The vertical schematic in Figure 4A should be explained in detail in its entirety. As it stands, no explanation is provided either in the figure caption or in the text. In particular, what does "elasticity driven" refer to?

(ii) The top snapshot in Figure 4A is labeled t_sim = 0 ns. Does it mean that the snapshot shown is the only chain configuration that the authors used to start the simulation, and that the snapshot does NOT represent the result of any time evolution, no matter how short the duration is? However, if that is the case, why is this snapshot identified with spinodal decomposition if it is not the product of a time evolution from a more homogeneous configuration?

(iii) Related to (ii) - do the rectangular boxes shown represent the entire simulation box or just part of the box containing the polymer chains? One would imagine that if the top snapshot represents spinodal decomposition, the simulation would have been started at a more uniform distribution a short time prior? Why is this not the case?

(iv) What precisely do the small yellow beads and black-colored springs in the zoomin image of Figure 4E represent?

We thank the reviewer for all these inspiring comments and questions. We agree that the original Figure 4 schematic did not sufficiently explain the sequence of dynamical events and the meaning of several graphical elements. We have therefore revised both the Figure 4 caption and the Results text to make the schematic self-contained and to clarify how it relates to the quantitative analyses in Figure 4F and G.

First, we replaced the phrase "elasticity driven" with a more precise description of "viscoelastic resistance". In the revised text, interfacial tension is described as favoring domain fusion thermodynamically, whereas transient inter-chain network connectivity within dense domains generates viscoelastic resistance to the deformation required for coalescence. This revision avoids implying that elasticity is the driving force and instead identifies it as a resistance that delays domain fusion kinetically during the plateau regime.

Second, we clarified the meaning of t_sim = 0 ns and its relation to spinodal decomposition. The system was equilibrated at a supercritical temperature to obtain a homogeneous one-phase state and was then instantaneously quenched to the target temperature. The label t_sim = 0 ns denotes the first snapshot immediately after the quench. It is not the only initial configuration used in all simulations; the reported kinetic metrics were averaged over six independent slab simulation replicas.

The t_sim = 0 ns snapshot is therefore described as a homogeneous but thermodynamically unstable post-quench state. Spinodal decomposition refers to the subsequent amplification of the initial density fluctuations after the quench, including the development of interconnected density fluctuations within 1-2 ns, rather than to a preceding evolution represented by the t_sim = 0 ns snapshot.

Third, we clarified that the rectangular snapshots in Figure 4A show the entire simulation box viewed along the z-axis. The subsequent snapshots show how post-quench density fluctuations grow and reorganize into dense domains during spinodal decomposition.

Finally, we clarified the symbols in the zoom-in schematic of Figure 4E. Yellow beads now denote residues involved in transient inter-chain contacts, and black springs denote schematic network connections formed by these contacts. These elements are meant to illustrate transient network connectivity and are not additional simulated particles or force-field terms.

Corresponding changes:

(1) (page 10, Figure 4A) The vertical schematic now labels the progression as "Thermodynamic instability", "Kinetic arrest (viscoelastic resistance)", "Domain coarsening (interfacial-tension dominated)", and "Dynamic equilibrium (chain self-diffusion)".

(2) (page 10, Figure 4E) The plateau schematic labels the competing effects as "Interfacial Tension" and "Transient network resistance".

(3) (page 10, Figure 4 caption) The caption defines the snapshots and the vertical schematic:

"Upper snapshots show the simulation box along the z-axis at tsim = 0, 10, and 500 ns. The vertical schematic summarizes the dynamical progression described in the main text, from the post-quench spinodal instability to kinetic arrest, domain coarsening, and dynamic equilibrium."

(4) (page 11, lines 368–370) The first recorded time point after the quench is defined explicitly:

"Here, tsim = 0 ns denotes the first snapshot immediately after the temperature quench, corresponding to a homogeneous but thermodynamically unstable nonequilibrium state."

(5) (page 10, Figure 4 caption) The yellow beads and black springs are defined:

"In the zoom-in view, yellow beads denote residues involved in transient inter-chain contacts, and black springs denote schematic network connections formed by these contacts."

(6) (page 12, lines 401–404) The Results explain the physical meaning of the transient network:

"In the zoom-in schematic in Figure 4E, this transient network is represented by connections between residues involved in inter-chain contacts, illustrating how multivalent interactions can resist domain deformation during the plateau regime."

(10) In discussing dynamic effects, it is useful to draw connections to related works on the effect of chain flexibility on "aging" of condensate [Biswas & Potoyan (2024) PRX 45:9222-9245 (https://journals.aps.org/prxlife/abstract/10.1103/PRXLife.2.023011)] and characterization of viscoelasticity in simulations of biomolecular condensates [Tejedor et al. (2023) J Phys Chem B 127:4441-4459 (https://pubs.acs.org/doi/10.1021/acs.jpcb.3c01292)], as the effects of desolvation can be explored further based on these prior works.

We thank the reviewer for these important references. We have added connections to simulation studies of condensate viscoelasticity and aging. The revised manuscript now places our dynamic results in the context of transient network connectivity, chain flexibility, sticker lifetime, desolvation-associated rigidification, and viscoelastic or aging-like material behavior.

We present these connections conservatively as relevant context and as future directions for extending the current model, rather than claiming a new universal dynamic mechanism.

Corresponding changes:

(1) (page 12, lines 397–399) The dynamics section cites simulation-based rheological analyses of condensate viscoelasticity:

"Similar viscoelastic effects have recently been quantified in molecular simulations of biomolecular condensates using rheological analyses of time-dependent material properties Tejedor et al. (2023)."

(2) (page 12, lines 399–401) The manuscript connects condensate aging to chain flexibility, sticker lifetime, and desolvation-associated rigidification:

"Molecular simulations of condensate aging have further highlighted the roles of chain flexibility, sticker lifetime, and desolvation-associated rigidification in promoting more solid-like states Biswas and Potoyan (2024)."

(3) (page 12, lines 423–427) The kinetic interpretation is connected to viscoelastic andaging-like behavior:

"The sensitivity of kinetic arrest and coarsening dynamics to desolvation parameters underscores the importance of incorporating desolvation features into coarse-grained potentials for more physically plausible molecular simulations of LLPS, especially when connecting microscopic interaction lifetimes to emergent viscoelastic or ageing-like material behavior."

(4) (page 16, lines 569–572) The Discussion identifies simulation-based rheological analysis as a future direction:

"An additional direction would be to combine these potentials with simulation-based rheological analyses to quantify how desolvation reshapes condensate viscoelasticity, aging-like maturation, and long-time material relaxation Tejedor et al. (2023); Biswas and Potoyan (2024)."

(11) Much of the present study is based on the original HPS formulation of Dignon et al. (2018). In this regard and also in anticipation of future development of improved interaction schemes, several issues should be stated and discussed, even if briefly:

(i) The original HPS model has a basic shortcoming in accounting for the relative interaction strengths of, among others, arginine vs lysine residues [Das et al. (2020) PNAS 117:28795-28805 (https://www.pnas.org/doi/10.1073/pnas.2008122117)].

(ii) Compared to 210-parameter pairwise interaction schemes, such as KH in Dignon et al. (2018) and Joseph et al. (2021), the 20-parameter interaction scheme is likely too restrictive to account for pairwise amino acid residue interactions [Wessén et al. (2022) J Phys Chem B 45:9222-9245 (https://pubs.acs.org/doi/10.1021/acs.jpcb.2c06181)].

(iii) The height of the desolvation barrier may vary significantly for different amino acid residue pairs, see, e.g., Figure 11 of Cinar et al. (2019) mentioned above (and references therein). The authors should discuss these nuances in the revised version. They may also wish to take them into consideration in future investigations.

We thank the reviewer for the suggestion to clarify these limitations. We have revised the Discussion to acknowledge explicitly the limitations of the 20-parameter hydropathy-scale representation relative to more flexible 210-parameter pairwise interaction schemes for describing amino-acid-pair interactions. We have also added discussion emphasizing that future desolvation-aware models should incorporate residue-pair-specific parameters for the desolvation barrier and solvent-separated potential well.

Corresponding changes:

(1) (page 13, lines 445–446) The scope of the averaged baseline parameterization is stated explicitly:

"This uniform parameterization captures the generic desolvation features of the PMFs but does not resolve residue-pair-specific variations in desolvation energetics."

(2) (page 15, lines 551–554) The Discussion identifies the limitation of the 20-parameter HPS representation, including Arg/Lys interactions:

"In particular, HPS-type models use a 20-parameter hydropathy-scale representation, which is useful for capturing generic IDP phase behavior but is not flexible enough to resolve residue-pair-specific chemical effects, such as the distinct interaction patterns of arginine and lysine residues Das et al. (2020)."

(3) (page 15, lines 554–557) The greater flexibility of 210-parameter pairwise schemes is described:

"More general 210-parameter pairwise interaction schemes, such as KH-type and related residue-pair-specific models, provide greater flexibility for encoding amino acid-pair preferences and capturing sequence-specific interaction heterogeneity Dignon et al. (2018b); Joseph et al. (2021); Wessén et al. (2022)."

(4) (page 15, lines 557–561) The limitation of using one averaged desolvation parameter set is stated:

"Second, the present desolvation model employs a single set of averaged parameters (αb, αss) for all residue pairs. While this simplification is effective for isolating the generic physical consequences of desolvation, it has limitations in describing the pair-specific variations in the desolvation barrier and the solvent-separated minimum."

(5) (page 15, lines 566–567; continues on page 16, lines 568–569) Residue-specific and temperature-dependent parameters are identified as a future extension:

"Future extensions of the model could therefore incorporate residue-specific and temperature-dependent desolvation parameters derived from bottom-up parameterization or expanded experimental datasets, thereby enhancing predictive accuracy for sequence-dependent LLPS."

Reviewer #2 (Public review):

Summary:

This manuscript addresses an important and timely question in the molecular simulation of biomolecular condensates. Most residue-level coarse-grained models used for IDP phase separation employ implicit solvent and represent effective interactions through relatively simple pairwise potentials. While these models have been very useful, they usually do not explicitly distinguish direct contacts from solvent-separated interactions, nor do they include an energetic barrier associated with water removal. This manuscript attempts to address that limitation by introducing desolvation-inspired terms into coarse-grained models and examining their consequences for phase behavior, chain conformations, dense-phase packing, and dynamics. Strengths:

The central idea is physically well motivated. Using a simple homopolymer model, the authors show that increasing the desolvation barrier suppresses phase separation, whereas stabilizing solvent-separated contacts enhances phase separation. They further show that solvent-separated interactions can reduce densephase over-compaction, which is a meaningful result given the known challenges in obtaining both accurate single-chain dimensions and realistic dense-phase properties from the same coarse-grained model. The finding that desolvation-like terms can reshape dense-phase packing without simply rescaling the overall interaction strength is interesting and could be useful for future model development. I also found the attempt to connect conformational changes across dilute and dense phases with thermal distance from the critical point to be intriguing. The dynamic analysis, including the FRAP-like simulations and the discussion of kinetic arrest during coarsening, adds another useful dimension to the work.

Weaknesses:

At the same time, there are several places where the manuscript would benefit from more careful framing. First, the desolvation terms are still effective coarse-grained parameters rather than a direct representation of water molecules. The language sometimes gives the impression that desolvation is being treated explicitly, whereas the model introduces desolvation-inspired effective interactions into an implicitsolvent framework.

We thank the reviewer for the positive assessment and constructive suggestions. We agree that the desolvation terms should be described as effective coarse-grained parameters rather than explicit water molecules. We have revised the manuscript to describe the model as a desolvation-aware implicit-solvent coarse-grained framework with desolvation-inspired effective interaction terms.

Corresponding changes:

(1) (page 1, lines 17–20) The Abstract identifies the model as an implicit-solvent CG model:

"Here, guided by all-atom simulations and experimental measurements, we develop a desolvation-aware implicit-solvent CG model by incorporating residue-level desolvation terms directly into the pairwise energy function and apply it to investigate LLPS of intrinsically disordered proteins."

(2) (page 4, lines 151–153) The Results describe the added contributions as desolvation-related effective terms:

"These observations underscore the importance of incorporating desolvation-related effective terms and exploring the effects of different desolvation strengths on the thermodynamics and kinetics of protein LLPS."

(3) (page 4, lines 172–174) The pair interaction is described as a desolvation-inspired effective potential:

"Together, these parameters shape the desolvation-inspired effective potential and modulate the statistical balance between direct-contact and solvent-separated configurations."

(4) (page 15, lines 541–543) The Discussion emphasizes that the framework retains water-mediated features within an implicit-solvent representation:

"By retaining key water-mediated features while preserving the computational efficiency of implicit-solvent representations, this framework provides a mechanistic means to decouple overall phase-separation propensity from condensed-phase packing."

Second, the conformational analysis is interesting, but the broader context of prior work on dilute-to-dense phase conformational reorganization of IDPs could be more clearly discussed. This would help clarify what is new in the present work, whether it is the conformational change itself, its dependence on desolvation terms, or the proposed scaling with distance from the critical point.

We thank the reviewer for this suggestion. We agree that the conformational change itself should be placed in the context of prior work. The contribution of the present analysis is not simply the observation that IDP conformations can reorganize upon condensation. Rather, we examine how desolvation-inspired effective terms modulate dilute- and dense-phase conformations and how the conformational change correlates with thermal distance from the critical point within the present model.

We have revised the Results section discussing Figure 3 to cite prior work and to state the interpretation of ΔR_g more clearly.

Corresponding changes:

(1) (page 6, lines 230–232) Prior work on conformational reorganization upon condensation is cited:

"Previous studies have shown that IDP condensation can reorganize chain conformations by redistributing the balance between intra-chain and inter-chain interactions Wei et al. (2017); Hazra and Levy (2021); Tesei et al. (2021); von Bülow et al. (2025)."

(2) (page 7, lines 241–242) The phase-dependent conformational response to the desolvation parameters is introduced:

"In addition to the difference between dilute- and dense-phase conformations, varying the desolvation parameters further reveals a phase-dependent conformational response (Figure 3A–C)."

(3) (page 7, lines 243–244) The stronger response in the dilute phase is stated directly:

"Increasing εb or decreasing εss shifts the dilute-phase Rg distributions toward larger values, whereas the dense-phase Rg remains comparatively insensitive to these parameter changes."

(4) (page 7, lines 247–249) The source of the desolvation-dependent variation in ΔR_g is identified:

"As a result, the desolvation-dependent variation in ΔRg = Rgdense − Rgdilute arises predominantly from the conformational changes of isolated chains in the dilute phase."

(5) (page 7, lines 254–256) The observed relationship is presented as an approximate trend in the simulated systems:

"Notably, data from the simulated systems approximately follow a common trend, revealing a strong correlation between the magnitude of conformational change and the thermal distance to the phase transition point (R2 = 0.942, Figure 3D)."

Third, the dynamic results are potentially useful, but the manuscript should more clearly articulate what is nontrivial beyond the expected slowing of local rearrangements by an added barrier in the potential.

Overall, I think this is a useful and potentially important contribution.

We thank the reviewer for this constructive comment and the positive overall assessment. We have revised the dynamics section to clarify that the nontrivial result lies in the competition between two effects: although the desolvation barrier directly slows local rearrangements, its reduction of dense-phase packing can reverse the net mobility trend at fixed temperature. At matched thermodynamic quench depth, the intrinsic slowing associated with energy-landscape roughness becomes evident. We also clarified that desolvation modulates transient kinetic arrest and domain-scale coarsening, not only local rearrangements.

Corresponding changes:

(1) (page 11, lines 353–357) The fixed-temperature and matched-quench-depth analyses are summarized as opposing contributions:

"Together, the fixed-temperature and renormalized analyses in Figure 4C and D reveal two distinct and opposing contributions of desolvation to condensate dynamics. At fixed temperature, increasing εb loosens dense-phase packing and thereby increases the measured diffusion coefficient, whereas at matched thermodynamic quench depth, the same parameter change suppresses chain mobility by roughening the microscopic energy landscape and slowing local rearrangements."

(2) (page 11, lines 358–361) The multiscale interpretation is stated explicitly:

"Condensate dynamics therefore emerge from a balance between density-regulated mobility and energy-landscape-regulated mobility, with macroscopic packing determining the dominant trend and microscopic barrier roughness imposing an additional kinetic modulation. This interplay highlights how desolvation reshapes condensate dynamics across multiple physical scales."

(3) (page 12, lines 421–423) The dynamics section distinguishes the result from simple local slowing:

"This picture shows that desolvation does more than slow down local chain rearrangements through an added barrier. It also regulates the balance between fluctuation growth, transient arrest, and domain coarsening, thereby shaping the evolution of phase-separated domains."

Reviewer #2 (Recommendations for the authors):

(1) The model is physically motivated and useful, but I would encourage the authors to be more precise in describing the added terms as desolvation-inspired effective interactions rather than explicit desolvation.

We thank the reviewer for the comment and suggestion. We have revised the manuscript accordingly and describe the added terms as desolvation-inspired effective interactions within an implicit-solvent CG framework throughout the Abstract, Results, and Discussion. More detailed changes are provided in our response to the first point raised in Reviewer #2's Public Review.

(2) The desolvation barrier is introduced as part of the equilibrium pair potential, and therefore it is expected to affect not only kinetics but also the phase boundary through changes in the configurational partition function. The manuscript would benefit from clarifying this point, since the term "barrier" may otherwise suggest a primarily kinetic role. In particular, the authors should explain whether the observed shift in T_c reflects a change in the effective pair attraction, for example, through the integrated Boltzmann weight or second virial coefficient, rather than only an entropic penalty associated with restricted configurations.

We thank the reviewer for this important point. We agree that the desolvation barrier is part of the equilibrium pair potential and therefore affects the phase boundary through the Boltzmann-weighted sampling of residue-pair configurations, not only through kinetic slowing.

Following this recommendation, we added a bead-level second virial coefficient analysis based on the effective pair potential. This analysis provides a pair-potentiallevel measure of the integrated effective attraction and clarifies why increasing the barrier lowers T_c, whereas stabilizing the solvent-separated minimum raises T_c.

Corresponding changes:

(1) (page 5, lines 203–205) The equilibrium role of the barrier is stated explicitly:

"At the pair-potential level, the desolvation barrier modifies the equilibrium Boltzmann weight and thereby alters the integrated effective attraction, as quantified by the bead-level second virial coefficient B2."

(2) (page 5, lines 205–207) The barrier-dependent second virial coefficient is connected to the shift in critical temperature:

"Specifically, increasing εb makes B2/σ3 larger (Figure 2—figure Supplement 1G), indicating a weaker integrated effective attraction and providing a thermodynamic basis for the lower Tc*."

(3) (page 5, lines 207–209) The solvent-separated-well trend is linked to a smaller second virial coefficient:

"By contrast, deepening the solvent-separated well εss elevates Tc* (Figure 2E), which is associated with the enhanced population of solvent-separated configurations and a smaller B2/σ3 (Figure 2—figure Supplement 1D, H)."

(4) (page 18, lines 662–665) The Methods specify the integration range and the quantity used to compare integrated effective attraction:

"The upper limit of integration rc is set as 3σ, which is sufficiently large to capture the full range of interactions while ensuring numerical convergence. The reduced value B2/σ3 was used to compare the integrated effective attraction under different desolvation parameters."

(5) (Figure 2—figure supplement 1G, H) The new panels report the integrated effective attraction:

"(G, H) Bead-level second virial coefficient (B2/σ3) calculated from the effective pair potential under varying εb at fixed εss = 0.02 kcal/mol (G) and varying εss at fixed εb = 3.12 cal/mol (H)."

(3) The conformational analysis in Figure 3 is interesting and potentially important. It would help to better place this result in the context of prior work showing dilute-todense phase conformational reorganization of IDPs, and to clarify what is new here beyond that broader observation.

We thank the reviewer for the comment and suggestion. We have revised the Results section discussing Figure 3 to place dilute-to-dense conformational reorganization of IDPs in the context of previous studies and then to emphasize the specific contribution of the present work.

The revised text clarifies that, within the present model, desolvation-inspired interactions mainly regulate chain conformations in the dilute phase, whereas dense-phase conformations remain comparatively insensitive. Detailed changes are provided in our response to the second point raised in Reviewer #2's Public Review.

(4) The proposed scaling between ΔR_g and distance from the critical point is intriguing, but the argument relies on simplifying assumptions. I would present this more as an empirical scaling supported by a plausible theoretical argument rather than a general result.

We thank the reviewer for this helpful suggestion. We agree that the correlation between Delta R_g and the distance from the critical point relies on simplifying assumptions and should not be presented as a general law. In the revised manuscript, we have softened the interpretation and now present this relationship as an empirical correlation supported by a simplified Flory-Huggins-based theoretical argument.

Corresponding changes:

(1) (page 7, lines 254–256) The relationship is described as an approximate trend in the simulated systems:

"Notably, data from the simulated systems approximately follow a common trend, revealing a strong correlation between the magnitude of conformational change and the thermal distance to the phase transition point (R2 = 0.942, Figure 3D)."

(2) (page 7, lines 256–258) The interpretation is limited to an association with thermal distance from the critical point:

"This result suggests that the conformational response to phase separation is closely associated with how far the system resides thermally from the critical point."

(3) (page 7, lines 263–264) The critical-condition derivation is shown explicitly:

"At the critical point, χ(Tc) = χc, which gives εeff = kBTcχc. Substituting this relation into the expression for χ(Tsim) yields χ(Tsim) = χcTc/Tsim."

(4) (page 7, lines 268–272) The structural relation is explicitly introduced as a first-order linear-response approximation:

"The thermodynamic driving force χ(Tsim) – χc can then be related to the structural observable ΔRg. Since ΔRg captures the structural transition from an intrachain-interaction-dominated state in the dilute phase to an interchain-interaction-dominated state in the dense phase, we assume, as a first-order approximation, that this conformational shift responds approximately linearly to the excess interaction strength, expressed as ΔRg ∝ [χ(Tsim) − χc]."

(5) (page 8, lines 285–287) The unscaled relation is labeled as an empirical scaling approximation:

"Although the complete relation in Equation (3) contains an additional Tsim factor, the unscaled quantities remain strongly correlated over the simulated range. We therefore use Tc − Tsim ∝ ΔRg as an empirical scaling approximation."

(5) The dynamics section would benefit from a statement of what is nontrivial, since a desolvation barrier is expected to slow local rearrangements.

We thank the reviewer for the comment and suggestions. As described above, we have revised the dynamics section to clarify what is nontrivial beyond the expected slowing of local rearrangements by an added barrier. The revised text emphasizes that desolvation affects condensate dynamics through competing effects of macroscopic packing and microscopic energy-landscape roughness, and that it also regulates transient kinetic arrest and domain-scale coarsening. More detailed changes are provided in our response to the third point raised in Reviewer #2's Public Review.

Version 2

Abstract

Biomolecular condensates play essential roles in cellular organization and are implicated in diverse pathological processes. Their formation is driven by liquid-liquid phase separation (LLPS), a process that requires coordinated multistep desolvation of biomolecular chains and multivalent inter-chain interactions. Although coarse-grained (CG) models with implicit solvent are widely used to probe LLPS thermodynamics and kinetics, they typically neglect water-mediated desolvation effects, limiting their accuracy and mechanistic interpretability. Here, guided by all-atom simulations and experimental measurements, we develop a desolvation-aware implicit-solvent CG model by incorporating residue-level desolvation terms directly into the pairwise energy function and apply it to investigate LLPS of intrinsically disordered proteins. Incorporating these desolvation interactions reshapes the phase diagram, alleviating dense-phase overcompaction. Notably, we observe an approximately linear correlation between the temperature gap (simulation temperature relative to the critical point) and the extent of conformational expansion accompanying the dilute-to-dense phase transition, a result further supported by theoretical analysis. We also find that desolvation barriers slow early density-fluctuation growth and shorten transient kinetic arrest, whereas solvent-separated contact interactions exert the opposite effects. Both terms further modulate chain mobility within mature condensates through competing packing and energy-landscape effects. Together, this framework enables an efficient representation of desolvation in CG simulations and reveals how desolvation energetics shape both the thermodynamic landscape and kinetic properties of biomolecular LLPS.

Introduction

Biomolecular condensates formed through liquid-liquid phase separation (LLPS) are increasingly recognized as fundamental regulators of diverse cellular processes, including stress responses Van Der Lee et al. (2014); Feric et al. (2016); Shin and Brangwynne (2017); Boeynaems et al. (2018); Burke et al. (2015); Biamonti and Vourc’h (2010); Riback et al. (2017), cellular signaling Wippich et al. (2013); Su et al. (2016), gene expression regulation Morimoto and Boerkoel (2013); Hnisz et al. (2017), and chromatin organization Strom et al. (2017). Dysregulation of condensate formation, in contrast, has been implicated in a wide range of human diseases, most notably cancer and numerous neurodegenerative disorders Molliex et al. (2015); Patel et al. (2015); Boeynaems et al. (2018); Conicella et al. (2016); Osterburg and Dötsch (2022). Intrinsically disordered proteins (IDPs), which lack stable tertiary structures and often contain multivalent interaction motifs, are key molecular drivers of LLPS Molliex et al. (2015); Uversky et al. (2015); Zhou et al. (2018); Dignon et al. (2020); Elbaum-Garfinkle et al. (2015). Understanding the molecular principles governing IDP-mediated LLPS is therefore essential for elucidating the mechanisms underlying both physiological condensate function and pathological phase transitions. A central yet incompletely understood aspect of LLPS is the multi-step desolvation of biomolecular chains as they transition from a homogeneously dispersed solution to a condensed phase. How desolvation energetics contribute to the thermodynamics and kinetics of LLPS remains a major open question.

Because experimentally characterizing the microscopic molecular events of LLPS at the sub-molecule level is still challenging for most experimental assays, a broad spectrum of computational approaches has been developed to characterize biomolecular phase separation. These range from mean-field Flory (1942); Huggins (1941) and random-phase-approximation polymer theories Lin et al. (2017); Lin and Chan (2017); McCarty et al. (2019) to diverse molecular simulation frameworks based on atomistic or coarse-grained (CG) force fields Dignon et al. (2018b,a); Davtyan et al. (2012); De Jong et al. (2013); Baul et al. (2019). Although all-atom molecular dynamics (MD) simulations can capture the microscopic processes of biomolecules with unprecedented temporal and spatial resolution, they become computationally prohibitive when applied to biomolecular phase separation involving a large number of protein chains. This challenge has motivated the development of CG models that reduce computational cost while retaining the essential physicochemical features governing phase behavior. These CG frameworks include early experiment-parameterized hydrophobicity-based models for IDPs Norgaard et al. (2008), the sticker-and-spacer paradigm and its lattice Monte-Carlo implementation LASSI Harmon et al. (2017); Ruff et al. (2021); Choi et al. (2019), the hydrophobicity-scale (HPS) family and its refinements Dignon et al. (2018b,a); Regy et al. (2021); Dannenhoffer-Lafage and Best (2021), and Mpipi, which explicitly incorporates cation-π interactions Joseph et al. (2021). Advanced multiscale models have also been pioneered to predict the coupled phase behavior of IDPs and chromatin with remarkable accuracy Espinosa et al. (2020). Higher-resolution or experimentally fine-tuned CG force fields (e.g., AWSEM-IDP Wu et al. (2018), Martini3 Zhang et al. (2023); Souza et al. (2021), COCOMO Valdes-Garcia et al. (2023); Jussupow et al. (2025), AICG2+ Li et al. (2014); Bian et al. (2024), and SIRAH Darré et al. (2015)) further capture hydrogen bonding, secondary-structure propensity, and sequence-dependent packing. Notably, the maximum entropy optimized force field (MOFF) integrates maximum entropy biasing and energy landscape theory, successfully predicting the experimental dimensions and macroscopic phase separation of both ordered and disordered proteins Latham and Zhang (2019, 2021). More recently, Tesei et al. introduced the CALVADOS model, which optimizes residue-specific interaction parameters within the HPS framework using SAXS and paramagnetic relaxation enhancement (PRE) data across a large IDP dataset Tesei et al. (2021); Tesei and Lindorff-Larsen (2022); Tesei et al. (2024), thereby improving the accuracy and transferability of CG models for predicting IDP conformations and LLPS behavior.

Despite these advances, most residue-level CG models rely on implicit solvent representations, in which individual water molecules are not explicitly represented. This simplification can neglect the contribution of water molecules in mediating inter-residue interactions, which can be crucial for the thermodynamics and kinetics of biomolecular folding and phase transitions Cheung et al. (2002); Liu and Chan (2005); Karanicolas and Brooks III (2002); Tarus et al. (2006); Wu et al. (2011b). In particular, the desolvation penalty associated with electrostatic and hydrophobic interactions exhibits marked differences between implicit and explicit solvent representations Salari and Chong (2012); Wu et al. (2011a). More importantly, conventional implicit-solvent CG models used for LLPS usually do not directly account for the multi-step desolvation process that accompanies the transition from a dilute solution to a dense condensate. Recent theoretical and experimental studies highlight that water reorganization and solvent-mediated entropy changes are not merely passive consequences of LLPS but active driving forces that shape condensate stability, internal structure, and dynamics Mukherjee and Schäfer (2023); Mukherjee et al. (2024); Li and Hou (2023). Related studies on hydrostatic pressure effects have further suggested that desolvation barriers and solvent-separated minima can help rationalize pressure-modulated LLPS behaviors, including the pressure-dependent reentrant phase separation of α-elastin Cinar et al. (2019, 2018). Nevertheless, implicit-solvent CG models remain widely used because they strike an effective balance between computational efficiency and residue-level resolution. Incorporating desolvation effects into such implicit-solvent models thus represents a particularly promising strategy for enhancing their physical realism without sacrificing tractability in LLPS simulations. Notably, residue-level desolvation terms have already been successfully employed in protein folding studies Cheung et al. (2002); Liu and Chan (2005); Ferguson et al. (2009); Chen and Chan (2014); Zhang and Chan (2010); Karanicolas and Brooks III (2002); Gasic and Cheung (2020); Camilloni et al. (2016). Within a structure-based modeling framework, Cheung et al. introduced a desolvation barrier separating the direct-contact minimum from the solvent-separated minimum in pairwise residue-residue potentials Cheung et al. (2002). Simulations demonstrated that this minimalist desolvation model captures water-expulsion events during the final stages of folding. Chan and colleagues further showed that incorporating a desolvation barrier markedly enhances folding cooperativity Liu and Chan (2005); Ferguson et al. (2009); Chen and Chan (2014); Zhang and Chan (2010). These previous successes provide a strong foundation for incorporating desolvation effects into coarse-grained models to more realistically investigate the critical contribution of desolvation to LLPS.

In this study, we advance the coarse-grained modeling of biomolecular LLPS by incorporating desolvation terms into the pairwise energy function of a CG model. By performing molecular simulations, we systematically investigate how desolvation influences phase behavior, thermodynamic properties, and the microscopic dynamics underlying condensate formation as well as the internal dynamics of the equilibrated dense phase. We further parameterize the desolvation terms using all-atom MD simulations and experimental Rg measurements, yielding a refined framework for investigating how generic desolvation features modulate the sequence-dependent phase behavior encoded by the underlying residue-level models.

Results

Effective representation of desolvation effects in coarse-grained models

In order to elucidate the role of water molecules in mediating inter-residue interactions, we first performed all-atom MD simulations with explicit solvent for a set of amino acid analogues and computed the potential of mean force (PMF) along inter-residue distances (r). Methane, methanol, acetate ion, ammonium ion, and methanamide were chosen to represent non-polar, polar, negatively charged, positively charged, and backbone-like residues, respectively. These amino acid analogues cover the key features of most amino acid types. The simulations were performed under ambient conditions (298 K and 1 atm), and the sampled snapshots were then analyzed to determine the inter-molecule distances and the corresponding PMF profiles. Additional simulation details are provided in the Methods section.

Figure 1A shows the PMF profile for the methane system. In addition to the deep potential well corresponding to the direct contact between two methane molecules (r ∼ 4 Å), there exists a shallow potential well at a larger distance (r ∼ 7 Å). This shallow minimum arises from transient, solvent-separated pseudo contacts between the solute residues, hereafter referred to as solvent-separated contact. To form a direct contact between the solutes, the water molecules mediating the solvent-separated contacts need to be excluded, which leads to a desolvation barrier between the two potential wells. Similar features have been reported in previous all-atom MD simulations Liu and Chan (2005). The PMF profiles extracted from the simulations for other amino acid analogues exhibit the same overall shape (Figure 1B and Figure 1—figure Supplement 1), but the positions of the direct-contact minimum (rdc), the solvent-separated minimum (rss), and the desolvation barrier (rb) vary depending on the analogue molecules. The depths of the two potential wells (ϵ and ϵss) and the height of the desolvation barrier (ϵb) likewise differ across analogue pairs.

Desolvation-related PMF features of intermolecular interactions.

(A) Potential of mean force along the intermolecular distances for the amino acid analogues from all-atom MD simulations with different solvent polarities. (B) Schematic diagram of desolvation effects. (C) Pairwise effective potential incorporating desolvation-inspired terms. Different curves correspond to different desolvation parameters.

The desolvation effect can also be modulated by solvent polarity and ionic conditions, which can be tuned experimentally by adding cosolvents and changing ionic strength, respectively. To examine the role of solvent polarity, we performed additional simulations of the methane system using water models with systematically varied polarity. For simplicity, the solvent polarity was tuned by uniformly scaling the partial charges of the hydrogen and oxygen atoms in the water molecules. This approach directly modulates the strength of solvent-solvent electrostatic interactions, which in turn dictates the stability of the hydration shell surrounding the solute. We compared the standard water model (1.0× charges) with reduced-polarity models (0.9× and 0.8× charges) (Figure 1A). The results reveal that a reduction in solvent polarity leads to a simultaneous decrease in the height of the desolvation barrier and the depth of the solvent-separated minimum. These observations underscore the importance of incorporating desolvation-related effective terms and exploring the effects of different desolvation strengths on the thermodynamics and kinetics of protein LLPS.

In this work, we introduce desolvation terms into the HPS model developed by Dignon et al. Dignon et al. (2018b). In this framework, each residue is simplified to a single spherical bead. The HPS energy function includes a bonded term and a nonbonded term (see Materials and methods). The nonbonded term is described by the short-range pairwise interactions and long-range electrostatic interactions. To represent the desolvation features, we added two Gaussian terms centered at rb and rss, corresponding to the desolvation barrier and the solvent-separated minimum, respectively. The resulting nonbonded effective potential is given by (1) U(r)=Φ(r,ϵ,λ,σ)+Gb(r,ϵb,rb,σb)−Gss(r,ϵss,rss,σss)=Φ(r,ϵ,λ,σ)+ϵbexp⁡[−(r−rb)22σb2]−ϵssexp⁡[−(r−rss)22σss2]. Here Φ(r, ϵ, λ, σ) denotes the nonbonded term in the original HPS model energy function Dignon et al. (2018b). The parameters λ and σ correspond to the hydrophobicity scale values and effective bead sizes, respectively, for the 20 amino acids. The widths of the Gaussian terms were fixed at σb = σss = 0.05 nm, while their centers were positioned at rb = rdc+0.15 nm and rss = rdc+0.30 nm. These offsets approximately correspond to the radius and diameter of a water molecule, respectively.

We first employed a homopolymer protein model with a chain length of 50 residues to illustrate the desolvation effect. For each residue, we use the parameter set extracted by averaging over the IDPs in the DisProt dataset Aspromonte et al. (2024), which gives an amino acid mass of mresidue = 110 amu (atomic mass unit), a hydrophobicity parameter λ = 0.640, and an effective bead size σ = 0.536 nm. The parameter ϵb controls the height of the desolvation barrier associated with the transition from solvent-separated to direct-contact configurations, whereas ϵss determines the depth of the local minimum corresponding to a solvent-separated contact. Together, these parameters shape the desolvation-inspired effective potential and modulate the statistical balance between direct-contact and solvent-separated configurations. By tuning the parameter values of ϵss and ϵb, we generate energy functions with varying desolvation contributions (Figure 1C), enabling systematic investigation of how desolvation influences the phase behavior and thermodynamics of IDPs. Because the effective interaction parameters used in the model are temperatureindependent, the absolute simulation temperature should not be directly interpreted as an experimental temperature. We therefore report all thermodynamic quantities, unless explicitly stated otherwise, in terms of a reduced temperature, defined as T ∗ = kBT /ϵ. Here, ϵ denotes the characteristic interaction energy scale of the system, which is set to ϵ = 0.2 kcal/mol following the convention of the HPS model Dignon et al. (2018b). Accordingly, T ∗ should be interpreted primarily as a model temperature that controls the relative strength of thermal fluctuations, rather than as having a direct quantitative correspondence with experimental temperature.

Thermodynamic impact of desolvation on the phase diagram

To capture the essential physics of IDP phase separation, we employed a coarse-grained model of 125 homopolymer chains using a slab simulation protocol (Methods). As a control, we first performed molecular dynamics simulations within the HPS model framework without considering the desolvation terms across a range of temperatures (Figure 2A–C). Representative snapshots clearly illustrate the transition from a homogeneous solution to a stable condensate as the temperature decreases (Figure 2B). We extracted time-averaged density profiles along the z-axis to identify coexisting dense and dilute phases and thereby constructed the binodal curve (Figure 2A, C), which reproduces phase behavior reported in previous molecular simulation studies Dignon et al. (2018b). We then repeated the slab simulations using desolvation-inspired effective potentials with varying ϵb and ϵss to investigate how desolvation reshapes the phase diagram and condensate structure.

Thermodynamic regulation and microscopic mechanisms of desolvation-mediated phase separation.

(A) Baseline phase diagram of the poly-50 system using the standard HPS model. (B) Representative simulation snapshots visualizing the transition from a stable condensate (T ∗ = 2.58) to a near-critical state (T ∗ = 2.98) and a homogeneous solution (T ∗ = 3.18). (C) Time-averaged density profiles along the z-axis identifying the coexisting dense and dilute phases. (D, E) Macroscopic phase boundaries under varying desolvation barrier heights ϵb (D) and solvent-separated potential depths ϵss (E). Insets show the monotonic dependence of Tc on the respective parameters. The lower panels schematically illustrate how changes in ϵb and ϵss alter the distribution of residue-pair configurations. The small circles indicate schematic populations of residue-pair configurations along the potential profile, with denser circles representing a higher population. (F, G) Renormalized phase behavior plotted against normalized temperature T ∗/Tc for varying ϵb (F) and varying ϵss (G).

To elucidate how desolvation regulates the thermodynamics of LLPS, we systematically mapped phase diagrams under varying desolvation parameters (Figure 2D and E). Increasing the desolvation barrier ϵb monotonically lowers the critical temperature Tc∗ (Figure 2D), suggesting a reduced phase separation propensity. Analysis of residue-residue radial distribution functions showed that higher ϵb suppresses the population of configurations near the barrier region (Figure 2—figure Supplement 1C). This reduction in the statistical weight of barrier-region configurations can be interpreted as an entropy-related configurational restriction and thus disfavors phase separation. At the pair-potential level, the desolvation barrier modifies the equilibrium Boltzmann weight and thereby alters the integrated effective attraction, as quantified by the bead-level second virial coefficient B2. Specifically, increasing ϵb makes B2/σ3 larger (Figure 2—figure Supplement 1G), indicating a weaker integrated effective attraction and providing a thermodynamic basis for the lower Tc∗. By contrast, deepening the solvent-separated well ϵss elevates Tc∗ (Figure 2E), which is associated with the enhanced population of solvent-separated configurations and a smaller B2/σ3 (Figure 2—figure Supplement 1D, H). This solvent-separated minimum provides effective free-energy stabilization for water-mediated configurations and thereby promotes phase separation. These opposing effects suggest that the desolvation potential regulates macroscopic phase behavior by redistributing residue-pair configurations between direct-contact, barrier-region, and solventseparated states (lower panels of Figure 2D and E).

Because the critical temperature is an intrinsic, sequence-specific property of a given system, it is also important to examine how desolvation strength alters the overall shape of the phase diagram when the temperature is normalized by the corresponding Tc∗. To do so, we rescaled the temperature axis by Tc∗, which effectively normalizes overall energetic strength and isolates desolvation-specific effects on the phase boundary. For the desolvation models with different barrier heights ϵb, the normalized binodal curves nearly overlap (Figure 2F). By contrast, varying the solvent-separated well depth ϵss results in distinct deviations in the density of the condensed phase (Figure 2G). Specifically, stronger water-mediated interactions lower equilibrium density by biasing the ensemble toward configurations with larger inter-residue separations (Figure 2—figure Supplement 1E, F). Thus, solvent-separated interactions directly regulate molecular packing and, by introducing solvent-separated configurations, alleviate the overcompaction typically encountered in coarse-grained models with implicit solvent. These results suggest that incorporating desolvation effects may enable the model to better capture the physicochemical structure of biological condensates.

Conformational modulation by desolvation in dense and dilute phases

We next investigated how desolvation modulates the conformational distribution of IDPs across dilute and condensed phases. Previous studies have shown that IDP condensation can reorganize chain conformations by redistributing the balance between intra-chain and inter-chain interactions Wei et al. (2017); Hazra and Levy (2021); Tesei et al. (2021); von Bülow et al. (2025). To quantify global chain dimensions, we calculated the radius of gyration (Rg) as a primary structural metric. Consistent with this picture, the resulting Rg distributions show that protein chains adopt more compact conformations in the dilute phase, whereas chains in the condensed phase exhibit more extended conformations (Figure 3B and C). This conformational expansion upon condensation reflects a transition from intra-chain-dominated interactions in the dilute phase to inter-chaindominated interactions in the dense phase Wohl et al. (2025). Within the condensed phase, protein chains preferentially form inter-chain contacts that lower the free energy of the system, compensating for the entropic cost associated with reduced translational and conformational freedom.

Effect of desolvation on protein conformations.

(A) Schematic illustration of the conformational distributions of the protein in the high- and low-density phases under different desolvation parameters. (B–C) Distribution of Rg with different ϵb (B) and ϵss (C) at Tsim∗=1.59. The solid and dashed lines represent the results for the condensed and dilute phases, respectively. The inset illustrates the mean value of Rg as a function of desolvation parameters. (D) Correlation between temperature difference (Tc∗−Tsim∗) and the averaged Rg difference between two phases. The purple and yellow dashed lines represent linear fits to the data obtained at low- (Tsim∗<1.2) and high-simulation-temperature (Tsim∗>2.4) regimes, respectively. The inset displays the improved linearity obtained when the temperature difference is rescaled by the simulation temperature, (Tc∗−Tsim∗)/Tsim∗. (E) Correlation between the density difference and Rg difference in the two phases with varying temperatures and desolvation parameters. The plot is shown on a log–log scale.

In addition to the difference between diluteand dense-phase conformations, varying the desolvation parameters further reveals a phase-dependent conformational response (Figure 3A–C). Increasing ϵb or decreasing ϵss shifts the dilute-phase Rg distributions toward larger values, whereas the dense-phase Rg remains comparatively insensitive to these parameter changes. This contrast suggests that desolvation more strongly affects chain dimensions in the dilute phase, while conformations in the dense phase are less responsive, likely reflecting the constraints imposed by intermolecular packing in the condensate. As a result, the desolvation-dependent variation in ΔRg=Rgdense−Rgdilute arises predominantly from the conformational changes of isolated chains in the dilute phase. Physically, ΔRg captures the transition from intramolecular-dominated collapse in the dilute phase to intermolecular-mediated expansion in the dense phase, thus serving as a metric for the relative strength of inter-chain versus intra-chain interactions.

To further elucidate the thermodynamic implications of desolvation-modulated conformations, we analyzed the relationship between the conformational shift ΔRg and the thermal distance to the critical point, defined as Tc∗−Tsim∗. Notably, data from the simulated systems approximately follow a common trend, revealing a strong correlation between the magnitude of conformational change and the thermal distance to the phase transition point (R2 = 0.942, Figure 3D). This result suggests that the conformational response to phase separation is closely associated with how far the system resides thermally from the critical point.

We rationalize this observation using the Flory-Huggins theory Huggins (1942); Flory (1942) as a bridge between the macroscopic thermal distance and the microscopic interaction strength. In this theoretical framework, the thermodynamic state is described by the interaction parameter χ = ϵeff /(kBT), where ϵeff represents the effective pairwise energy scale specific to each desolvation model Flory (1953). At the critical point, χ(Tc) = χc, which gives ϵeff = kBTcχc. Substituting this relation into the expression for χ(Tsim) yields χ(Tsim) = χcTc/Tsim. Subtracting χc from both sides gives (2) χ(Tsim)−χc=χcTc−TsimTsim. For systems with the same chain length, χc is a fixed constant. Thus, the deviation from the critical interaction parameter is directly related to the rescaled thermal distance (Tc − Tsim)/Tsim.

The thermodynamic driving force χ(Tsim) − χc can then be related to the structural observable ΔRg. Since ΔRg captures the structural transition from an intrachain-interaction-dominated state in the dilute phase to an interchain-interaction-dominated state in the dense phase, we assume, as a first-order approximation, that this conformational shift responds approximately linearly to the excess interaction strength, expressed as ΔRg ∝ [χ(Tsim) − χc]. Combining these relations gives (3) Tc−Tsim∝TsimΔRg. This relation explains the approximate linearity observed in Figure 3D, while also accounting for the systematic deviations caused by the temperature factor Tsim, which is evident from the varying slopes at different simulation temperatures. To test this theoretical argument and the underlying linear-response assumption, we replotted the data against the rescaled thermal distance (Tc∗−Tsim∗)/Tsim∗. As shown in the inset of Figure 3D, this normalization yields a stronger linear relationship, supporting the above argument that ΔRg scales with the excess interaction strength χ − χc, and further suggesting that the conformational response is closely associated with the rescaled thermal distance from the critical point.

We further show the relationship between the phase density difference (Δρ = ρdense − ρdilute) and the conformational difference (ΔRg) between coexisting phases in Figure 3E. Based on the empirical correlation between ΔRg and the thermal distance from the critical point, one can relate the density contrast Δρ to ΔRg through the critical-scaling relation by Δρ = A(Tc − T)β (Equation (7)), where β is the critical exponent. Although the complete relation in Equation (3) contains an additional Tsim factor, the unscaled quantities remain strongly correlated over the simulated range. We therefore use Tc −Tsim ∝ ΔRg as an empirical scaling approximation. Using this approximation in the critical-scaling relation yields Δρ ∝ (ΔRg)β . On logarithmic scales, a clear power-law dependence emerges (Figure 3E), with Δρ∝ΔRg0.4, which is consistent with our theoretical expectation. The fitted exponent (≈ 0.4) is in reasonable agreement with the theoretical critical exponent β = 0.325 based on the 3D Ising model Rowlinson and Widom (2013). Collectively, these results reveal a fundamental coupling between microscopic conformational reorganization and macroscopic packing density during biomolecular condensation, highlighting conformational reorganization as a structural signature of the phase transition.

Dynamic consequences of desolvation in phase separation

Beyond governing thermodynamic phase boundaries and conformational distribution, the desolvation potential fundamentally reshapes the dynamical landscape of IDP condensates. The interplay between inter-chain interactions and solvent exclusion not only dictates the stability of the dense phase but also modulates the transport properties within the condensate and the macroscopic kinetics of phase separation. To probe chain mobility within the dense phase, we employed a strategy analogous to fluorescence recovery after photobleaching (FRAP) experiments following prior molecular dynamics simulation work Yamada and Takada (2023), tracking the spatiotemporal evolution of specific chains initially localized at the slab center (Figure 4A and B). The systems were compared at the same simulation temperature Tsim∗. For the standard HPS model without desolvation, the tagged chains achieve a uniform distribution that matches the equilibrium density profile within 50 ns (Figure 4B), consistent with the liquid-like nature of the condensate. For the representative desolvation model, the spatial distribution of tracked chains remains biased toward the initial center position after 50 ns and approaches equilibrium on a prolonged timescale (∼ 150 ns). These contrasting relaxation profiles demonstrate that desolvation alters chain mobility within the condensate.

Desolvation-mediated modulation of diffusion and coarsening dynamics.

(A) Snapshots of phase-separation dynamics after a temperature quench and the subsequent FRAP-like analysis of chain self-diffusion in an equilibrated slab. Upper snapshots show the simulation box along the z-axis at tsim = 0, 10, and 500 ns. The vertical schematic summarizes the dynamical progression described in the main text, from the post-quench spinodal instability to kinetic arrest, domain coarsening, and dynamic equilibrium. Lower panels show enlarged slab views for tracking chains initially located near the slab center, with the TAMSD plot quantifying their mobility. (B) Density profiles along the z-axis for the entire system (orange) and for the highlighted chains (red) at different time lags, with and without desolvation. Shaded histograms show instantaneous snapshots, and solid curves represent normalized time averages over 1 μs. (C) Diffusion coefficients as a function of the desolvation strength (ϵb, ϵss) and dense-phase density (ρdense) at different temperatures. (D) Reduced diffusion coefficient K~ versus quench depth for varying ϵb. The schematic insets summarize the reduced chain mobility observed for higher desolvation barriers at comparable quench depths. (E) Schematic illustration of the three-stage phase-separation mechanism, including early density-fluctuation growth, a transiently arrested plateau, and late-stage coarsening. In the zoom-in view, yellow beads denote residues involved in transient inter-chain contacts, and black springs denote schematic network connections formed by these contacts. (F, G) Kinetics of density fluctuations and domain growth under varying ϵb (F) and ϵss (G). Top panels show the time evolution of the normalized density variance ⟨σ^ρ2⟩, with the inset showing the characteristic time τ1/2. Bottom panels show the growth of the average slab size ⟨ξ⟩ on a logarithmic scale. The colored timeline highlights the initial fluctuation, plateau, and late-stage coarsening regimes.

To more quantitatively assess the influence of desolvation on chain mobility, we computed the time-averaged mean squared displacement (TAMSD) Metzler et al. (2014) for chains within the slab. The chain motion exhibits sub-diffusive scaling, TAMSD = K ⋅ Δtα, with a characteristic exponent α ≈ 0.91 (Figure 4A, bottom right panel). This deviation from normal diffusion (α = 1) likely reflects the combined effects of molecular crowding, multivalent dynamic interaction networks, and finite-size confinement within the dense phase. At extended lag times, the TAMSD curves naturally plateau as they approach the geometric limit imposed by the slab thickness. Notably, the observed exponent is comparable to simulation results reported for protein diffusion in threedimensional condensates (α ≈ 0.85) Watanabe et al. (2025), consistently highlighting the restricted nature of chain dynamics within the dense phase. Given the relatively consistent values of α across different interaction parameters, we use K as a quantitative measure of the chain diffusion rate.

The diffusion prefactor K exhibits a clear dependence on the interaction parameters (Figure 4C). At a fixed temperature, K increases with the desolvation barrier ϵb and decreases with the solventseparated minimum depth ϵss (Figure 4C, left panels). These opposing trends can be largely understood from the distinct effects of the two desolvation parameters on condensate packing. As illustrated in Figure 4C (right panel), K is predominantly correlated with the bulk density (ρdense), with further modulation by temperature. As expected, higher-temperature systems generally exhibit slightly larger K values at comparable densities. Consistent with the density trends established in Figure 2D and E, a higher ϵss promotes tighter packing and increases the effective friction, whereas a higher ϵb prevents close contacts and creates greater free volume that facilitates chain motion. These results indicate that, at a fixed absolute temperature, changes in dense-phase packing serve as the primary determinant of chain mobility.

However, this density-dominated view is incomplete because it convolutes intrinsic energy-landscape features with thermodynamic state shifts. Since variations in ϵb and ϵss significantly alter the critical temperature, comparisons made at a constant simulation temperature inherently involve systems positioned at different thermodynamic quench depths, defined as the normalized temperature deviation from the critical point. This highlights the distinct thermodynamic driving forces for different systems. To reduce this thermodynamic-state dependence, we performed a more carefully controlled comparison. First, we scaled the simulation temperature by the critical temperature of each system and used Tsim∗/Tc∗ as the control parameter for quench depth, with a lower value corresponding to a deeper quench. This provides more comparable thermodynamic conditions across the different interaction parameters. Second, to account for the leading linear temperature dependence associated with Langevin dynamics, we defined a reduced diffusion prefactor K~=K/T as detailed in the Methods section. Under these controlled conditions, a clearer parameter-dependent kinetic trend emerges (Figure 4D). At the same quench depth, a higher desolvation barrier ϵb (darker blue points) lowers K~, consistent with a rougher energy landscape that hinders the configurational rearrangements required for chain diffusion. This kinetic divergence becomes more pronounced at deeper quenches but largely vanishes near the critical point, indicating that the kinetic influence of the desolvation barrier depends on the thermodynamic state. Notably, the standard HPS model exhibits much faster temperature-normalized mobility (Figure 4D, red triangles), suggesting that omitting desolvation energetics may lead to an overestimation of chain mobility due to reduced roughness of the effective energy landscape.

Together, the fixed-temperature and renormalized analyses in Figure 4C and D reveal two distinct and opposing contributions of desolvation to condensate dynamics. At fixed temperature, increasing ϵb loosens dense-phase packing and thereby increases the measured diffusion coefficient, whereas at matched thermodynamic quench depth, the same parameter change suppresses chain mobility by roughening the microscopic energy landscape and slowing local rearrangements. Condensate dynamics therefore emerge from a balance between density-regulated mobility and energy-landscape-regulated mobility, with macroscopic packing determining the dominant trend and microscopic barrier roughness imposing an additional kinetic modulation. This interplay highlights how desolvation reshapes condensate dynamics across multiple physical scales.

Desolvation-modulated kinetic arrest of coarsening dynamics

Elucidating the molecular events during the early stage of phase separation is challenging experimentally. We therefore examined droplet-growth dynamics at the early stage of phase separation by performing molecular simulations with desolvation-aware effective potentials. Initially, the system was equilibrated at a supercritical temperature (T ∗ = 2.98) for 10 ns to ensure a homogeneous distribution. The temperature was then quenched to T ∗ = 1.79, located deep within the two-phase coexisting region, to initiate spontaneous phase separation (Figure 4A). Here, tsim = 0 ns denotes the first snapshot immediately after the temperature quench, corresponding to a homogeneous but thermodynamically unstable non-equilibrium state. During the subsequent time evolution, the initially uniform distribution rapidly develops interconnected density fluctuations within 1 ns to 2 ns, which then break up into discrete high-density domains along the z-axis. Over time, these domains gradually coalesce into larger slabs, exhibiting the characteristic morphological evolution of spinodal decomposition Cahn and Hilliard (1958).

To quantify the phase separation kinetics, we adopted a two-stage metric strategy. In the early stage, where interfaces are diffusive, we monitored the normalized density variance (⟨σ^ρ2⟩, see Supplementary Information) derived from the instantaneous density profile ρ(z) to track the growth rate of density fluctuations. As clear interfaces emerged, we switched to tracking the average slab size (⟨ξ⟩), defining dense domains as continuous regions where ρ(z) exceeds a threshold value (here the average system density was used as the threshold). To ensure statistical robustness, all kinetic metrics were averaged over six independent slab simulation replicas. The resulting kinetic profiles (Figure 4F and G) resolve a three-stage dynamical pathway. In the first stage, immediately after the temperature quench, the system undergoes a rapid exponential growth in density variance (top panels). This behavior arises from the growth of initial density fluctuations in a thermodynamically unstable state and is characteristic of spinodal decomposition Cahn (1961). Stronger effective attraction, corresponding to higher ϵss or lower ϵb, accelerates these initial fluctuations, as quantified by the monotonic decrease in the characteristic time scale τ1/2, defined as the time required to reach 50% of the maximum normalized variance.

Intriguingly, after dense domains have formed, the system enters an intermediate plateau regime in which domain growth is temporarily arrested before late-stage coarsening begins. This kinetic arrest is evident from the slab-size evolution in Figure 4F and G, where ⟨ξ⟩ remains nearly unchanged for a distinct period, and is visually corroborated by the density profiles at 5 ns in Figure 4E, where distinct domains are observed to form but fail to fuse immediately. This behavior is consistent with the characteristics of viscoelastic phase separation Tanaka (2000). In this regime, interfacial tension favors domain fusion, whereas transient inter-chain connectivity within the dense domains generates viscoelastic resistance to the shape deformation required for coalescence, thereby delaying domain fusion. Similar viscoelastic effects have recently been quantified in molecular simulations of biomolecular condensates using rheological analyses of time-dependent material properties Tejedor et al. (2023). Molecular simulations of condensate aging have further highlighted the roles of chain flexibility, sticker lifetime, and desolvation-associated rigidification in promoting more solid-like states Biswas and Potoyan (2024). In the zoom-in schematic in Figure 4E, this transient network is represented by connections between residues involved in inter-chain contacts, illustrating how multivalent interactions can resist domain deformation during the plateau regime. Quantitative analysis in Figure 4F and G further reveals that the duration of this plateau regime depends strongly on the desolvation parameters. Systems with higher ϵss or lower ϵb exhibit a longer plateau, indicating that stronger inter-chain attractions reinforce the transient network and prolong kinetic arrest. This trend corroborates our viscoelastic interpretation, as enhanced intermolecular connectivity increases the resistance to domain deformation and fusion. This kinetic trap is particularly pronounced in the standard HPS model (red trajectories in Figure 4F and G), consistent with its relatively stronger pairwise attractions.

At later times (t > 10 ns), the system leaves the plateau regime and enters a coarsening-and- growth stage, where domain fusion proceeds through interfacial-tension-driven relaxation and Brownian aggregation. The average slab size follows a power-law scaling ⟨ξ⟩ ∼ tα, with α ≈ 0.26 in our simulations. This exponent is slightly lower than the asymptotic prediction of Brownian motion coalescence (BMC), α = 1/3 Majumdar et al. (1994), likely because the viscoelastic nature of the polymer-rich phase and the sub-diffusive motion of individual chains (TAMSD ∼ Δt0.91) constrain macroscopic domain diffusion Doi et al. (1988).

Collectively, these analyses reveal a desolvation-modulated kinetic pathway that links early density fluctuations, transient domain arrest, and late-stage coarsening into a unified dynamical picture, as summarized by the vertical schematic in Figure 4A and the mechanistic illustration in Figure 4E. This picture shows that desolvation does more than slow down local chain rearrangements through an added barrier. It also regulates the balance between fluctuation growth, transient arrest, and domain coarsening, thereby shaping the evolution of phase-separated domains. The sensitivity of kinetic arrest and coarsening dynamics to desolvation parameters underscores the importance of incorporating desolvation features into coarse-grained potentials for more physically plausible molecular simulations of LLPS, especially when connecting microscopic interaction lifetimes to emergent viscoelastic or aging-like material behavior. Beyond modulating LLPS thermodynamics and conformational ensembles as discussed above, desolvation therefore acts as a critical regulator of phase-separation dynamics.

Parameterization of desolvation model for IDPs

The results from the homopolymer model discussed above underscore the critical role of desolvation in regulating both the thermodynamics and dynamics of LLPS. Consequently, incorporating water-mediated interactions into residue-level coarse-grained models is essential for capturing the physics of protein phase separation more realistically. To this end, we developed a transferable parameterization strategy that combines information derived from all-atom simulations with calibration against experimental observables of IDPs.

Directly determining residue-specific desolvation parameters, specifically the barrier height ϵb and the solvent-separated minimum depth ϵss, remains challenging. To bridge this gap, we leveraged all-atom molecular dynamics simulations of amino acid analogues. Using the global interaction energy scale ϵ of the underlying HPS-type potential as a reference, we expressed the desolvation amplitudes as ϵb = αbϵ and ϵss = αssϵ. By fitting the PMF profiles obtained from all-atom simulations (Figure 1—figure Supplement 1), we obtained coefficients of comparable magnitude across the limited set of analogue pairs examined. For simplicity and transferability, we therefore adopted the averaged values αb = 0.33 and αss = 0.06 as a global baseline for our coarse-grained model. This uniform parameterization captures the generic desolvation features of the PMFs but does not resolve residue-pair-specific variations in desolvation energetics.

To evaluate the effectiveness of these physics-derived parameters, we first integrated them into the framework of the standard HPS model. While the HPS model performs remarkably well in capturing sequence-dependent properties of LLPS, previous studies have noted that parameters optimized solely against single-chain Rg may have limited transferability to the condensed phase, resulting in an overestimation of phase-separation propensity, as reflected by higher critical temperatures (Tc) Dignon et al. (2018b). We assessed the model performance on a dataset of 47 IDPs taken from Tesei et al. Tesei et al. (2021), which was subsequently used to calibrate the CALVADOS2desolvation parameters. The deviation of simulated single-chain Rg from experimental SAXS data was quantified by χRg2=[(Rg,exp−Rg,sim)/σexp]2, where σexp represents the experimental uncertainty. When the desolvation parameters derived from all-atom simulations (αb = 0.33, αss = 0.06) were added to the HPS model without modifying its original ϵ, we observed a substantial improvement in conformational accuracy. The average discrepancy ⟨χRg2⟩ decreased drastically from 1,082.27 (original HPS) to 85.88 (HPS-desolvation) (Figure 5B). This improvement indicates that introducing the desolvation terms effectively mitigates the over-compaction tendency of the original HPS model, likely because of competition between direct and water-separated contacts.

Parameterization of desolvation terms for the HPS and CALVADOS2 models based on IDPs.

(A) Schematic workflow of the desolvation parameterization. (B) Correlation between experimental Rg and simulation Rg for the original HPS model (blue) and the revised HPS model with default desolvation scales (αb = 0.33 and αss = 0.06) (purple). The ⟨χRg2⟩ for different models are also shown. (C) Correlation between experimental Rg and simulation Rg for the original CALVADOS2 model (orange), the revised CALVADOS2 model with default desolvation (yellow), and the revised CALVADOS2 model with optimized desolvation (αb = 0.3, αss = 0.03 and ϵ = 0.262 kcal/mol, red). (D) Coexistence curves of FUS LC simulated with the original HPS model (blue), the revised HPS model with default desolvation scales (purple), the energy-rescaled HPS model (green), the default CALVADOS2 model (orange), and the revised CALVADOS2 model with optimized desolvation scales (red). The green shaded regions highlight the deviations of the desolvation models relative to the original frameworks.

We also incorporated the desolvation terms into CALVADOS2 Tesei and Lindorff-Larsen (2022), a force field whose hydrophobicity parameters (λ) were reparameterized using Bayesian optimization to reproduce experimental Rg and PRE data. Directly adding desolvation terms to CALVADOS2 led to substantial chain over-expansion (⟨χRg2⟩=643.30), reflecting the fact that CALVADOS2 is already finely tuned to reproduce experimental data through a top-down optimization procedure and the inclusion of additional desolvation terms is incompatible with its parameterization. To address this, we implemented a systematic optimization strategy in which the desolvation barrier height (ϵb), solvent-separated minimum depth (ϵss), and the overall energy factor ϵ were scanned over broad parameter ranges to identify the parameter set that minimizes deviations between simulated and experimental Rg values. The full optimization workflow is summarized in the flowchart in Figure 5A, and the resulting error landscapes are shown in Figure 5—figure Supplement 1 and Figure 5 figure Supplement 2 . The optimized parameter set (Table 1) reduced ⟨χRg2⟩ to 9.54 on the calibration dataset (Figure 5C), yielding agreement with experimental Rg values comparable to that of the original CALVADOS2 model.

After parameterization of the energy scale, we further evaluated the calibrated model by probing the phase behavior of the FUS low-complexity domain (FUS LC) as a test case. Experimental observations indicate that the critical temperature is around 25 °C Burke et al. (2015). While the original HPS model yields a critical temperature of Tc ∼ 340 K, incorporating the desolvation terms into the HPS framework (HPS-desolvation) lowers the model Tc to ∼ 290 K (Figure 5D), shifting it toward the experimentally observed room-temperature regime. It is worth noting that the residue-level coarse-grained models used here employ temperature-independent effective interaction parameters. As a result, the temperature values reported here cannot be interpreted as quantitatively equivalent to experimental temperatures, particularly when they deviate substantially from room-temperature conditions. Beyond this reduction in the excessive phase-separation propensity, we observed that the desolvation model mitigates the overestimation of condensate density common in implicit solvent coarse-grained simulations Regy et al. (2021); Benayad et al. (2020). In the CALVADOS2 framework, the inclusion of desolvation reduces the dense phase density while maintaining an overall phase boundary comparable to the original model (Figure 5D). Similarly, for the HPS model, when we rescaled the interaction strength ϵ to match the single-chain Rg and thereby align the phase boundary with the HPS-desolvation system (Figure 5—figure Supplement 3), we observed the same behavior. A similar trend was observed in the simulations of the LAF-1 RGG domain, where CALVADOS2-desolvation preserved a comparable coexistence curve while modestly reducing the dense-phase density relative to the original CALVADOS2 model (Figure 5—figure Supplement 4). Quantitatively, across both frameworks, the desolvation-modulated models consistently yield a packing density approximately 5% lower than the original models. This distinction underscores a limitation of standard Lennard-Jones potentials where a single parameter ϵ simultaneously controls chain dimension, overall phase-separation propensity, and condensate density. In contrast, our model introduces tunable parameters ϵb and ϵss that provide an additional physical handle for modulating dense-phase packing without simply rescaling the overall interaction strength. The observed density reduction can be attributed to an increased population of solvent-separated configurations relative to direct contacts. This effect provides an effective representation of hydration-associated packing in protein condensates, thereby offering a more physically realistic description of IDP phase separation.

Discussion and Conclusion

Coarse-grained molecular simulations have become indispensable for elucidating the molecular principles underlying LLPS, complementing both theoretical frameworks and experimental observations. While residue-level implicit-solvent models have achieved remarkable success in capturing sequence-dependent phase behavior, they usually do not separately account for desolvation effects associated with water-mediated residue-residue association. As a result, they often face a trade-off when simultaneously describing single-chain dimensions, phase-separation propensity, and dense-phase packing. Specifically, interaction parameters calibrated to match single-chain dimensions or phase-separation propensity frequently result in an overestimation of the condensed phase density. This limitation suggests that standard Lennard-Jones potentials, which couple overall phase-separation propensity and dense-phase packing through a single energetic parameter, may lack the physical dimensionality required to fully describe the thermodynamics of hydrated protein condensates. The present results therefore suggest that desolvation is not merely a microscopic correction to residue-residue contacts, but an additional physically motivated component in coarse-grained interaction models that can modulate LLPS thermodynamics, dense-phase packing, conformational reorganization, and condensate dynamics in partially separable ways.

To address this challenge, we developed a desolvation-aware implicit-solvent CG framework that incorporates desolvation barrier and solvent-separated terms into the pairwise potential. Application of this model to homopolymer systems demonstrates that desolvation substantially reshapes phase behavior. Incorporating desolvation modulates the critical temperature and alters the phase diagram by regulating intermolecular packing within the dense phase. Moreover, the model reveals a tight coupling between desolvation-controlled chain compaction during the formation of the condensate phase and the thermal distance from criticality (temperature gap), providing a theoretical basis for linking single-chain conformational changes (ΔRg) to macroscopic phase behavior. Dynamic analysis further captures the three characteristic stages of early LLPS, including spinodal decomposition, kinetic arrest, and coarsening. Crucially, it reveals how the desolvation effect regulates the kinetics of these stages by modulating the effective inter-chain interactions and the roughness of the energy landscape. These results underscore the importance of accounting for desolvation energetics in governing both the thermodynamics and kinetics of LLPS. These findings may also provide a useful physical basis for future studies of pressure-dependent condensate behavior, as pressure-induced changes in hydration, solvent-separated states, and desolvation barriers have been proposed to contribute to pressure-modulated and reentrant LLPS Dias and Chan (2014); Cinar et al. (2019, 2018).

Motivated by these insights, we further developed a transferable parameterization strategy for residue-level models. The resulting desolvation-aware CG model improves the accuracy of IDP conformational ensembles and LLPS thermodynamics while preserving the computational efficiency inherent to CG methodologies. By retaining key water-mediated features while preserving the computational efficiency of implicit-solvent representations, this framework provides a mechanistic means to decouple overall phase-separation propensity from condensed-phase packing. As illustrated in Figure 5D, incorporating desolvation into CALVADOS2 reduces ρdense while preserving an overall phase boundary comparable to that of the original model. This decoupling effectively mitigates the overcompaction of the dense phase commonly observed in implicit solvent simulations with CG models, thereby making the simulations qualitatively more consistent with the physical nature of biological condensates.

Despite these advances, several limitations remain in the present implementation. First, the current model introduces desolvation terms on top of residue-level CG frameworks with simplified interaction parameterizations. In particular, HPS-type models use a 20-parameter hydropathy-scale representation, which is useful for capturing generic IDP phase behavior but is not flexible enough to resolve residue-pair-specific chemical effects, such as the distinct interaction patterns of arginine and lysine residues Das et al. (2020). More general 210-parameter pairwise interaction schemes, such as KH-type and related residue-pair-specific models, provide greater flexibility for encoding amino-acid-pair preferences and capturing sequence-specific interaction heterogeneity Dignon et al. (2018b); Joseph et al. (2021); Wessén et al. (2022). Second, the present desolvation model employs a single set of averaged parameters (αb, αss) for all residue pairs. While this simplification is effective for isolating the generic physical consequences of desolvation, it has limitations in describing the pair-specific variations in the desolvation barrier and the solvent-separated minimum. Explicit-solvent PMF analyses have shown that desolvation barrier heights and solvent-separated minima can differ substantially among residue pairs and may also exhibit temperature dependence Cinar et al. (2019); Debiec et al. (2014); MacCallum et al. (2007). In addition, effective interactions themselves can be temperature dependent, as demonstrated in analytical theories and coarse-grained simulations of biomolecular condensates Lin et al. (2017); Dignon et al. (2019); Chakravarti and Joseph (2025). Future extensions of the model could therefore incorporate residue-specific and temperature-dependent desolvation parameters derived from bottom-up parameterization or expanded experimental datasets, thereby enhancing predictive accuracy for sequence-dependent LLPS. An additional direction would be to combine these potentials with simulation-based rheological analyses to quantify how desolvation reshapes condensate viscoelasticity, aging-like maturation, and long-time material relaxation Tejedor et al. (2023); Biswas and Potoyan (2024).

In summary, this work provides a physically interpretable route for incorporating solvent-mediated interaction features into implicit-solvent CG models, providing a basis for future residue-pair-specific and condition-dependent models of biomolecular condensates. By enabling an efficient representation of key desolvation-related features, the model provides a robust framework for simulating biomolecular condensates with improved thermodynamic, structural, and kinetic fidelity. Future developments, such as residue-pair-specific desolvation energetics and extensions to multicomponent systems involving nucleic acids, hold strong potential for deepening our mechanistic understanding of the complex phase behaviors that organize cellular biochemistry.

Methods

All-atom MD simulations

The all-atom MD simulations were performed using GROMACS. Ten amino acid analog molecules were solvated in a cubic water box under periodic boundary conditions. The OPLS-AA force field was used to model the solute molecules, and the TIP4P water model was employed to represent solvent. Na+ and Cl− ions were added to neutralize the system, yielding a total system size of 1,010 atoms. After an initial energy minimization of 5 × 104 steps, the system was equilibrated in the NVT ensemble at 298 K for 100 ps, followed by an NPT equilibration at 298 K and 1 atm for 1 ns. Temperature and pressure were controlled using the Nosé–Hoover thermostat and Parrinello– Rahman barostat Parrinello and Rahman (1981), respectively. A time step of 2 fs was used for all MD steps. The simulation details are provided in Table 2. Subsequently, a production simulation of 100 ns was performed under the same NPT conditions for structural and thermodynamic analysis. The intermolecular separation r between two amino acid analogues was defined as the distance between their C atoms. The potential of mean force (PMF) was computed as PMF(r) = −kBT ln P (r), where P (r) is the radial probability density obtained from the production trajectory.

Coarse-grained simulations

We utilized the slab method Dignon et al. (2018b) to extract the densities of the coexisting dilute and dense phases. In these simulations, 125 protein chains were placed in a rectangular box of dimensions 15 nm × 15 nm × 280 nm. The density profile along the z-axis was computed to identify the condensed and dilute regions, from which thermodynamic quantities such as the critical temperature were subsequently determined. Simulations were performed over a range of different temperatures. To incorporate the desolvation terms into the pairwise interaction potentials of the HPS model, we used the azplugins package (https://github.com/mphowardlab/azplugins) together with HOOMD-blue v2.9.7 Anderson et al. (2020). We also implemented the desolvation terms for both the HPS and CALVADOS models based on the OpenMM open-source toolkit Eastman et al. (2023). GPU-accelerated simulations were performed on NVIDIA GeForce RTX 4090D and Tesla V100 GPUs, and CPU-based simulations were carried out on Kunpeng 920B. All simulations used a time step of 0.01 ps. Each trajectory was first equilibrated for 3 × 107 steps, followed by a production run of 1 × 108 steps. Additional details of the simulation framework are provided in Table 3.

Hydrophobicity Scale Model

Our desolvation model was developed based on the hydrophobicity scale (HPS) coarse-grained model by Dignon et al. Dignon et al. (2018b), in which each residue is represented by a single bead. The potential energy comprised a bonded term, an electrostatic term and a short-range pairwise term. The bonded interaction term is given by a harmonic potential V b = kb(ri,i+1 − r0)2 with the spring constant kb = 10 kcal/mol/Å2 and the bonded length r0 = 3.8 Å. r i,i+1 represents the distance between the consecutive residues i and i + 1 along the protein chain. The electrostatic interaction term is modeled by a Coulombic potential with Debye-Hückel Debye and Hückel (1923) electrostatic screening caused by salt ion concentration. The function is given by (4) Vijele(r)=qiqj4πDrexp⁡(−r/κ), where κ is the screening length and D = 80, which is the relative dielectric constant of water. For the short-range pairwise interaction, the HPS model uses a 6-12 Lennard-Jones potential with the interaction strength given by hydrophobicity scale Kapcha and Rossky (2014), which can describe the effective interaction between residues. The hydrophobicity value λ, ranging from 0 to 1, varies for different residues. The arithmetic average is used for the interaction parameter between two residues, i.e. hydrophobicity value λi,j = (λi + λj)/2 and amino acid size σi,j = (σi + σj)/2. The energy function of the short-range pairwise interaction is given by (5) Φ(r)={ΦLJ+(1−λ)ϵ,ifr⩽21/6σλΦLJ,otherwise where ΦLJ is standard 6-12 Lennard-Jones potential (6) ΦLJ=4ϵ[(σr)12−(σr)6]. The free parameter ϵ is set as 0.2 kcal/mol, with which the molecular simulations can best reproduce the experimental results of Rg values for a set of IDPs.

Simulation Framework

We used a slab simulation scheme in the coexistence simulations. Initially, 125 chains were randomly distributed in a cubic box with periodic boundary conditions at 200 K. The box length is first scaled to 15 nm, which can effectively prevent protein chains from interacting with their periodic images generated by periodic boundary conditions. Then, the x- and y- dimensions were held fixed and the z-dimension is extended to 280 nm. The temperature is then gradually increased to the target temperature and the simulations are conducted at different temperatures for 1 μs in the NVT ensemble maintained by a Langevin thermostat, with a mass-normalized damping rate of 0.01 ps−1. A time step of 0.01 ps was used for all the simulations. We used HOOMD-Blue v2.9.7 packages Anderson et al. (2020) and OpenMM v8.2.0 Eastman et al. (2023) to conduct the simulations.

The slab density profile along the z-axis is then determined after the system had equilibrated. In order to avoid the influence of periodic boundary conditions, we first move the center of mass of the system along the z-axis direction to the position of z = 0. We then determined the distribution of dense phase and dilute phase of the phase separated system using the defined thresholds. The area with density higher than 0.95 of the maximum density (ρmax) is treated as dense phase, and their average density is regarded as the phase density of the dense phase (ρdense). For the dilute phase, the threshold is taken as ρmin + 50 mg/mL, where ρmin is the minimum density. The density of dilute phase (ρdilute) is then determined by the same method. If ρmax − ρmin < 50 mg/mL, the system is considered not to undergo phase separation.

Critical Temperature

The critical temperature was determined based on the vapor-liquid interfacial properties of Lennard-Jones chains Blas et al. (2008). The critical temperature Tc was obtained from ρdense and ρdilute at different temperatures by (7) ρdense−ρdilute=A(Tc−T)β, where β is the critical exponent, which was set to 0.325 Rowlinson and Widom (2013) and A is the fitting parameter. Data with temperature below the critical temperature was used to fit this equation. The phase density at critical temperature was determined by the law of rectilinear diameters Davies (1912) (8) ρdense+ρdilute2=B+C⋅T, with B and C being the free parameters determined from the simulation data. At the critical temperature, the distinction between dense phase and dilute phase disappears, which means ρdense = ρdilute = ρc. The phase density was then calculated by ρc = B + C ⋅ Tc. The phase diagram was then constructed from the dense- and dilute-phase densities as functions of temperature.

Bead-level second virial coefficient

The bead-level second virial coefficient was calculated from the effective pair potential as (9) B2=2π∫0rc[1−exp⁡(−U(r)kBT)]r2dr, where U (r) is the desolvation-aware effective pair potential defined in Equation (1). The upper limit of integration rc is set as 3σ, which is sufficiently large to capture the full range of interactions while ensuring numerical convergence. The reduced value B2/σ3 was used to compare the integrated effective attraction under different desolvation parameters. All B2 values reported in Figure 2—figure Supplement 1G, H were evaluated at the same reduced temperature. This bead-level quantity provides a pair-potential-level measure of effective attraction.

Temperature Normalization of the Generalized Diffusion Coefficient

To investigate the intrinsic influence of the interaction potential on chain dynamics, we must isolate the structural contributions from the trivial thermal acceleration. While keeping the quench depth (T /Tc) constant provides a comparison at matched quench depth, it necessitates varying the absolute temperature T, which inherently alters the base diffusion rate via thermal fluctuations. To resolve this, we adopt the constant quench depth approach to ensure thermodynamic consistency and introduce a normalized metric to factor out the leading linear thermal contribution.

Motivated by the Einstein relation for Langevin diffusion, we describe the leading temperature dependence of the generalized diffusion coefficient K in terms of the thermal energy kBT and the effective friction coefficient ζ : (10) K(T)∝kBTζ(T,ϕ,ϵ). This approximation is motivated by the weakly subdiffusive dynamics observed here, for which α ≈ 0.91 and remains nearly constant across the simulated conditions. Here, the effective friction coefficient ζ encapsulates the resistance to chain motion. Following the hybrid theoretical framework proposed by Macedo and Litovitz Macedo and Litovitz (1965), which combines free-volume theory with activated rate theory, we phenomenologically factorize the effective friction into packing-dependent and energy-barrier-dependent contributions: (11) ζ(T,ϕ,ϵ)∝exp⁡(γv∗vf(ϕ))×exp⁡(Eeff(ϵ)kBT) where the first term represents the hindrance due to limited free volume (vf), governed by the packing density ϕ, and the second term is the Arrhenius factor describing the activated hopping over an effective energy barrier Eeff, which corresponds to the roughness of the energy landscape induced by the desolvation potential.

In our analysis, by fixing the quench depth (T /Tc), we expect the macroscopic packing density ϕ to remain approximately comparable across different systems (as verified by the phase diagram in main text). Consequently, the free-volume contribution to friction is effectively invariant. Substituting this back into Equation (10), the diffusion coefficient scales as: (12) K(T)∼T⋅exp⁡(−Eeff(ϵ)kBT) The prefactor T represents the leading linear thermal acceleration inherent to Langevin dynamics. To better isolate the intrinsic energy-landscape signature, we define the reduced diffusion coefficient K~ by normalizing out this explicit linear thermal contribution: (13) K~≡KT∝exp⁡(−Eeff(ϵ)kBT) Thus, K~ serves as a physically motivated metric to better isolate the intrinsic influence of the energy function on chain dynamics. By normalizing out the explicit linear thermal contribution inherent to Langevin dynamics, this definition helps decouple the specific kinetic modulation imposed by the desolvation potential from the background thermal acceleration and thermodynamic driving forces. Consequently, K~ allows for a more direct assessment of how the interaction potential alters transport properties after removing the leading linear temperature effect required to maintain thermodynamic consistency.

Calculation of Normalized Density Variance

To quantify the rate of density fluctuation growth during the early stage of phase separation, we first computed the instantaneous spatial variance of the density profile along the z-axis, σρ2(t), defined as (14) σρ2(t)=1Lz∫0Lz[ρ(z,t)−ρ¯]2dz, where ρ(z, t) is the local density at position z and time t, and ρ¯ is the mean density of the system. To enable a fair comparison of kinetics across systems with distinct interaction parameters (ϵss and ϵb), it is crucial to decouple the rate of phase separation from the difference in equilibrium dense-phase densities. Therefore, we normalized the instantaneous variance by its equilibrium baseline: (15) σ^ρ2(t)=σρ2(t)⟨σρ2⟩eq. Here, the normalization factor ⟨σρ2⟩eq is the time-averaged density variance calculated from the final 30% of the simulation process. This period corresponds to the steady state characterized by the formation of a well-defined slab with stabilized phase densities. By using this normalized metric σ^ρ2(t), we effectively factor out the variations in final density contrast, allowing the observed differences in the time-evolution curves to more directly reflect the acceleration or deceleration of the kinetic process itself.

Desolvation Parameters

Because residue-pair-specific experimental desolvation data were unavailable, the empirical parameters αb and αss were estimated using all-atom molecular dynamics simulations of amino acid analogues with explicit solvent. The potential of mean force (PMF) obtained from the simulation results is shown in Figure 1—figure Supplement 1. Methane (CH4), methanol (CH3OH), acetate ion (C2H3O2−), ammonium ion (NH4+), and methanamide (CH3NO) were chosen as representative amino acid analogues for non-polar, polar, negatively charged, positively charged, and backbone-like residues, respectively. For each analogue pair, αb and αss were obtained by fitting the non-bonded energy function with desolvation effects (Equation (1)) to the simulation-derived PMF. The λ values assigned to the analogues were taken from the corresponding amino acids. Specifically, we used λGLY = 0.649 for methane, λSER = 0.595 for methanol and λCYS = 0.595 for methanamide. The results are shown in Table 1.